---
title: "Cache Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cache Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The DSP chunks need a signal-processing companion. Everything else here runs
# on the bundled demo corpus, so this vignette works offline.
eval_dsp <- requireNamespace("superassp", quietly = TRUE)
```

## What reindeer caches

When you call `quantify(..., .use_cache = TRUE)` or
`enrich(corpus, ..., .use_cache = TRUE)`, results are written to a
persistent cache in `<corpus>/.quantify_cache/`. The next time you run
the same call with the same parameters, reindeer reads the result from
disk instead of re-running the DSP.

Two companion packages keep their own caches and provide their own
management tools:

- **erodex** — simulation result tables. See `erodex::list_simulations()`.
- **protoscribe** — draft annotation cache. See `protoscribe`'s docs.

This vignette covers only the quantify/enrich cache, which reindeer
manages itself.

## Set up a corpus

Everything below runs against the corpus that ships with the package:

```{r}
library(reindeer)

corp <- demo_corpus()
segs <- collect(query(corp, "Phoneme =~ .+"))
nrow(segs)
```

## Inspect what's cached

```{r}
inspect_cache(corp)
```

`inspect_cache()` prints a one-line summary (file count and total size)
and returns a tibble of cache entries. It is the recommended entry point.

## Tell hits from misses

A successful cache lookup means `quantify()` didn't redo the work. You
can see this at row level through the `.cache_status` column:

```{r, eval = eval_dsp}
formants <- quantify(segs, superassp::trk_rms, .at = 0.5, .use_cache = TRUE)
table(formants$.cache_status)
```

On a second identical call every row is served from disk:

```{r, eval = eval_dsp}
formants_again <- quantify(segs, superassp::trk_rms, .at = 0.5, .use_cache = TRUE)
table(formants_again$.cache_status)
```

If a repeat call still reports `"miss"`, the cache key changed — most
often because you changed `.at`, a DSP parameter passed through `...`, or
the speaker's `Age` / `Gender`, all of which are part of the key.

## Clean up old files

```{r}
manage_cache(corp)                                   # status
manage_cache(corp, action = "list")                  # inventory
manage_cache(corp, action = "clean", days_old = 30)  # dry run
```

Always preview with the default `dry_run = TRUE` before you actually
delete: pass `dry_run = FALSE` once you are happy with what the preview
lists.

## Size thresholds

For long-running pipelines you may want a hard cap:

```{r}
check_cache_size(file.path(corp@basePath, ".quantify_cache"),
                 warn_threshold = "500 MB",
                 max_threshold  = "2 GB")
```

This warns above `warn_threshold` and warns more strongly above
`max_threshold` — useful inside scripts where you want to stop before
the cache fills the disk.

## Cache invalidation in practice

Three changes cause a re-compute on the next call:

- The signal file changed (different mtime).
- You changed a DSP parameter passed through `...`.
- You changed the `Age` / `Gender` metadata for that bundle; those feed
  `dsp_parameters()`, which is part of the cache key.

For a clean start, remove `<corpus>/.quantify_cache/` and re-run.

## Where next

- `?inspect_cache`, `?manage_cache`, `?check_cache_size`.
- `vignette("getting_started")` — where `.use_cache = TRUE` first appears
  in the workflow.
