---
title: "Cache Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Cache Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

## What reindeer caches

When you call `quantify(..., .use_cache = TRUE)` or
`enrich(corpus, ..., .use_cache = TRUE)`, the results are written to a
persistent cache in `<corpus>/.quantify_cache/`. The next time you
run the same call with the same parameters reindeer reads the result
from disk instead of re-running the DSP.

Two companion packages keep their own caches and provide their own
management tools:

- **erodex** — simulation result tables. See `erodex::list_simulations()`.
- **protoscribe** — draft annotation cache. See `protoscribe`'s docs.

This vignette covers only the quantify/enrich cache, which reindeer
manages itself.

## Inspect what's cached

```{r}
library(reindeer)

corp <- corpus("path/to/ae_emuDB")
inspect_cache(corp)
```

`inspect_cache()` prints a one-line summary (file count and total
size) and returns a tibble sorted by file size. It is the recommended
entry point.

## Tell hits from misses

A successful cache lookup means `quantify()` didn't redo the work.
You can see this at row level via the `.cache_status` column:

```{r}
formants <- quantify(segs, superassp::forest, .use_cache = TRUE)

table(formants$.cache_status)
# "hit"  -> read from cache
# "miss" -> just computed and stored
```

On a second identical call you should see every row marked `"hit"`.
If you don't, the cache key changed — most often because you changed
`.at`, `nominalF1`, or the Age/Gender of the speaker.

## Clean up old files

```{r}
manage_cache(corp)                                  # status
manage_cache(corp, action = "list")                 # inventory
manage_cache(corp, action = "clean", days_old = 30) # dry-run preview
manage_cache(corp, action = "clean", days_old = 30, dry_run = FALSE)
```

Always preview with the default `dry_run = TRUE` before you actually
delete.

## Size thresholds

For long-running pipelines you may want a hard cap:

```{r}
check_cache_size(corp@.cache_dir,
                 warn_threshold = "500 MB",
                 max_threshold  = "2 GB")
```

This emits a warning above `warn_threshold` and a stronger one above
`max_threshold` — useful inside scripts where you want to bail out
before the cache eats the disk.

## Cache invalidation in practice

Three changes will cause a re-compute on next call:

- The signal file changed (different mtime).
- You changed any DSP parameter passed through `...`.
- You changed the Age/Gender metadata for that bundle (they feed
  `dsp_parameters()`, which is part of the cache key).

If you ever want a clean start, delete `<corpus>/.quantify_cache/`
and re-run the pipeline.

## Where next

- `?inspect_cache`, `?manage_cache`, `?check_cache_size`.
- `vignette("getting_started")` — where `.use_cache = TRUE` first
  appears in the workflow.
