---
title: "End-to-end pipeline with reindeer + companion packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end pipeline with reindeer + companion packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = identical(Sys.getenv("REINDEER_EVAL_VIGNETTES"), "true"))
```

## What this vignette covers

A full pipeline that walks a corpus from raw recordings to publishable
analysis, gluing reindeer's verbs to four companion packages:

- **superassp** for the audio-side DSP (formants, F0, intensity, ...).
- **eggstract** for the electroglottography (EGG) DSP when EGG signal
  files are present.
- **protoscribe** for draft annotation generation (VAD, VOT, periods,
  MOMEL/INTSINT).
- **erodex** for parameter-grid simulations.

reindeer itself supplies the corpus class, metadata system, EQL query
engine, lazy pipeline, provenance accounting, tidyverse-friendly
verbs, and FAIR documentation emitters. The companion packages plug
into well-defined extension points and are gated by
`requireNamespace()` checks so the pipeline degrades gracefully when a
companion is not installed.

## 1. Load the corpus

```{r}
library(reindeer)
corp <- corpus("path/to/your_emuDB")
corp
```

## 2. Document baseline metadata

Set the database-level project descriptor and any per-bundle flags
that gate downstream tooling. Setting `HasEGG = TRUE` on the bundles
that have EGG recordings tells `quantify_egg()` which segments to
process.

```{r}
set_metadata(corp, list(
  project = list(
    name = "Demo Corpus",
    description = "End-to-end pipeline demo for the reindeer R package."
  )
))

# Per-bundle: declare EGG availability.
for (bndl in c("bundle_001", "bundle_002")) {
  set_metadata(corp, list(HasEGG = TRUE),
               session = "ses_01", bundle = bndl)
}

# Per-session: speaker biographics that drive age/gender-aware DSP.
# These flow automatically into quantify()/enrich() when
# .use_metadata = TRUE — formant ranges, pitch limits, and window sizes
# all get picked from the speaker's row of the DSPP table.
set_metadata(corp,
             list(Speaker = "P001", Age = 25, Gender = "Female"),
             session = "ses_01")

# Preview what each bundle's DSP parameters will be:
dsp_parameters(corpus_obj = corp)
```

## 3. Suggest annotations with protoscribe

protoscribe's `draft_*` functions generate annotation suggestions
without writing them to the corpus. reindeer wraps them so review can
happen in the EMU-webApp via `serve_app()`.

```{r}
# VAD: voice-activity-driven Segment-level suggestions.
vad_suggestions <- propose_annotations(corp, type = "vad",
                                        .review = FALSE)
head(vad_suggestions)

# Open the review UI; accept/edit interactively, save when happy.
propose_annotations(corp, type = "periods", .review = TRUE)
```

## 4. Query and navigate

```{r}
vowels <- query(corp, "Phonetic == V") |>
  ascend_to("Word") |>
  dplyr::filter(end - start > 80)
provenance(vowels)
dropped_rows(vowels)
```

## 5. Quantify the audio side (superassp)

`.use_metadata = TRUE` (the default) tells `quantify()` to pick the
formant ranges and window size from the speaker's `Age` / `Gender`
metadata.

```{r}
audio_measures <- vowels |>
  collect() |>
  quantify(superassp::forest, .at = seq(0, 1, 0.1))
```

### Iterating with the cache

When you tweak downstream code and re-run the pipeline, turn the cache
on. The first call populates it; every subsequent call with the same
parameters and metadata returns rows from disk. The `.cache_status`
column lets you tell hits from misses at a glance.

```{r}
# First pass: cache populates (.cache_status all "miss")
audio_measures <- vowels |>
  collect() |>
  quantify(superassp::forest, .at = seq(0, 1, 0.1), .use_cache = TRUE)
table(audio_measures$.cache_status)

# Iterating: same call returns from cache (.cache_status all "hit")
audio_measures <- vowels |>
  collect() |>
  quantify(superassp::forest, .at = seq(0, 1, 0.1), .use_cache = TRUE)
table(audio_measures$.cache_status)
```

Editing a bundle's `Age` or `Gender` invalidates only that bundle's
rows; the rest of the cache stays warm.

## 6. Quantify the EGG side (eggstract)

`quantify_egg()` only runs on bundles whose `HasEGG = TRUE` flag is
set; the rest pass through untouched.

```{r}
egg_measures <- vowels |>
  collect() |>
  enrich(corp, with = "metadata") |>  # joins HasEGG and speaker metadata
  quantify_egg(.using = eggstract::ksvF0,
                .at = seq(0, 1, 0.1))
```

## 7. Reshape for plots

```{r}
long_formants <- audio_measures |> pivot_tracks_longer()

# All formants on one panel
ggplot2::autoplot(audio_measures, type = "formants")

# Or build it from primitives
library(ggplot2)
ggplot(long_formants) + geom_formant_trajectory()
```

## 8. Parameter-grid simulation (erodex)

When you need to sweep DSP parameters, hand off to erodex. The
simulation result re-enters the reindeer pipeline as a regular
extended_segment_list.

```{r}
library(erodex)
erodex::quantify_simulate(
  vowels |> collect(),
  .using = superassp::forest,
  .simulate = list(nominalF1 = seq(500, 900, 100)),
  .simulation_store = "simulations/formants"
)
```

## 9. Export to field-standard formats

Interop with Praat and ELAN is one line each:

```{r}
write_textgrid(vowels |> collect(), "out/vowels.TextGrid")
write_eaf(vowels |> collect(), "out/vowels.eaf",
          media_url = "audio/clip.wav")
```

## 10. Publishable corpus documentation

`describe_corpus()` emits five FAIR artifacts in one call: README,
CMDI, DataCite, CITATION.cff, and a schema.org JSON-LD `Dataset`
descriptor.

```{r}
describe_corpus(corp)
list.files(corp@basePath, pattern = "CITATION|cmdi|jsonld|datacite|README")
```

## What's next

- **Lazy optimizer** (`explain()`, filter pushdown, lazy
  metadata enrichment) — slated for v1.1.
- **DOI minting** — Zenodo / DataCite integration on top of
  `describe_corpus()` — also v1.1.
- **BIDS-speech export** — when the spec firms up.
