---
title: "End-to-end pipeline with reindeer + companion packages"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{End-to-end pipeline with reindeer + companion packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Each section runs only when its companion package is installed, so the
# vignette renders anywhere and shows real output wherever it can.
has_egg   <- requireNamespace("eggstract", quietly = TRUE)
has_dsp   <- requireNamespace("superassp", quietly = TRUE)
```

## What this vignette covers

A full pipeline that walks a corpus from raw recordings to publishable
analysis, gluing reindeer's verbs to two DSP companion packages:

- **superassp** for the audio-side DSP (formants, F0, intensity, ...).
- **eggstract** for the electroglottography (EGG) DSP when EGG signal
  files are present.

reindeer itself supplies the corpus class, metadata system, EQL query
engine, lazy pipeline, provenance accounting, tidyverse-friendly
verbs, and FAIR documentation emitters. These two DSP backends plug
into well-defined extension points and are gated by
`requireNamespace()` checks so the pipeline degrades gracefully when a
companion is not installed.

Draft annotation generation and DSP parameter-grid simulation are
fully independent optional add-ons — see
[protoscribe](https://github.com/humlab-speech/protoscribe)'s
`vignette("protoscribe-with-reindeer")` and
[erodex](https://github.com/humlab-speech/erodex)'s own vignette set.

## 1. Load the corpus

```{r load}
library(reindeer)
library(dplyr, warn.conflicts = FALSE)

corp <- demo_corpus()   # stands in for corpus("path/to/your_emuDB")
md <- get_metadata(corp)
ses <- unique(md$session)[1]
bndl <- md$bundle[md$session == ses][1]
```

## 2. Document baseline metadata

Set the database-level project descriptor and any per-bundle flags
that gate downstream tooling. Setting `HasEGG = TRUE` on the bundles
that have EGG recordings tells `quantify_egg()` which segments to
process.

```{r metadata}
set_metadata(corp, list(
  project = list(
    name = "Demo Corpus",
    description = "End-to-end pipeline demo for the reindeer R package."
  )
))

# Per-bundle: declare EGG availability for one recording.
set_metadata(corp, list(HasEGG = TRUE), session = ses, bundle = bndl)

# Per-session: speaker biographics that drive age/gender-aware DSP.
# These flow automatically into quantify() when
# .use_metadata = TRUE - formant ranges, pitch limits, and window sizes
# all get picked from the speaker's row of the DSPP table.
set_metadata(corp, list(Speaker = "P001", Age = 25, Gender = "Female"),
             session = ses)

# Preview what each bundle's DSP parameters will be:
head(dsp_parameters(corpus_obj = corp), 3)
```

## 3. Sequence and dominance queries (EQL)

Beyond simple label matches, EQL supports sequences (`->`) and
dominance (`^`) directly against the SQLite cache. Find vowels
immediately followed by a `t`, then the syllables that dominate each
`t`:

```{r eql}
vowel_then_t <- query(corp, "[Phonetic =~ [aeiou] -> Phonetic == t]") |>
  reindeer::collect()
head(vowel_then_t)

dominating_syllables <- query(corp, "[Syllable == S ^ Phonetic == t]") |>
  reindeer::collect()
head(dominating_syllables)
```

## 4. Query and navigate

```{r query}
vowels <- query(corp, "Phonetic =~ [aeiou]") |>
  filter(end - start > 30) |>      # drop the shortest gestures
  scout(steps_forward = 1)         # and keep what follows each one

provenance(vowels)
dropped_rows(vowels)
```

## 5. Quantify the audio side (superassp)

`.use_metadata = TRUE` (the default) tells `quantify()` to pick the
formant ranges and window size from the speaker's `Age` / `Gender`
metadata.

```{r quantify, eval = has_dsp}
audio_measures <- vowels |>
  quantify(superassp::trk_formant_forest, .at = seq(0, 1, 0.1)) |>
  reindeer::collect()

dim(audio_measures)
head(as.data.frame(audio_measures), 3)[, c("labels", "F1_Hz", "F2_Hz", ".time_point")]
```

```{r dsp-note, echo = FALSE, eval = !has_dsp}
cat("superassp is not installed here, so the audio-side sections are shown without output.\n")
```

### Iterating with the cache

When you tweak downstream code and re-run the pipeline, turn the cache
on. The first call populates it; every subsequent call with the same
parameters and metadata returns rows from disk. The `.cache_status`
column lets you tell hits from misses at a glance.

```{r cache, eval = has_dsp}
audio_measures <- vowels |>
  quantify(superassp::trk_formant_forest, .at = seq(0, 1, 0.1),
           .use_cache = TRUE) |>
  reindeer::collect()

table(audio_measures$.cache_status)
```

Editing a bundle's `Age` or `Gender` invalidates only that bundle's
rows; the rest of the cache stays warm.

## 6. Quantify the EGG side (eggstract)

`quantify_egg()` only runs on bundles whose `HasEGG = TRUE` flag is
set; the rest pass through untouched.

```{r egg, eval = has_egg}
egg_measures <- vowels |>
  biographize(corp) |>   # joins HasEGG and speaker metadata
  quantify_egg(.using = eggstract::ksvF0, .at = seq(0, 1, 0.1))
```

```{r egg-note, echo = FALSE, eval = !has_egg}
cat("eggstract is not installed here, so this section is not run.\n")
```

## 7. Reshape for plots

```{r reshape, eval = has_dsp, fig.width = 6, fig.height = 4}
long_formants <- audio_measures |> pivot_tracks_longer()
dim(long_formants)

ggplot2::autoplot(audio_measures, type = "formants")
```

```{r primitives, eval = has_dsp, fig.width = 6, fig.height = 4}
library(ggplot2)
ggplot(long_formants) + geom_formant_trajectory()
```

## 8. Export to field-standard formats

Interop with Praat and ELAN is one line each:

```{r export}
segs <- vowels |> reindeer::collect()
out <- file.path(tempdir(), "export")
dir.create(out, showWarnings = FALSE)

write_textgrid(segs, file.path(out, "vowels.TextGrid"))
write_eaf(segs, file.path(out, "vowels.eaf"), media_url = "audio/clip.wav")
list.files(out)
```

## 9. Publishable corpus documentation

`describe_corpus()` emits five FAIR artifacts in one call: README,
CMDI, DataCite, CITATION.cff, and a schema.org JSON-LD `Dataset`
descriptor.

```{r fair}
describe_corpus(corp)
list.files(corp@basePath, pattern = "CITATION|cmdi|jsonld|datacite|README")
```

## What's next

- **Lazy optimizer** (`explain()`, filter pushdown, lazy
  metadata enrichment) — slated for v1.1.
- **DOI minting** — Zenodo / DataCite integration on top of
  `describe_corpus()` — also v1.1.
- **BIDS-speech export** — when the spec firms up.
