---
title: "Getting Started with reindeer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with reindeer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

**reindeer** is an R package that extends [emuR](https://github.com/IPS-LMU/emuR) for working with speech data in challenging conditions. It provides:

- **Optimized workflows** for large speech corpora
- **Metadata management** with SQLite caching
- **Age/gender-specific** signal processing parameters
- **Intelligent caching** for computationally intensive operations
- **Modern S7 classes** for type safety and performance

This guide will walk you through the basics of using reindeer for your speech analysis workflow.

## Installation

```{r}
# Install from GitHub (development version)
# install.packages("remotes")
remotes::install_github("your-username/reindeer")

# Load the package
library(reindeer)
```

## Quick Start: 5-Minute Workflow

Here's a complete analysis workflow in just a few lines:

```{r}
# 1. Load your database
corp <- corpus("path/to/your_database_emuDB")

# 2. Query for segments
vowels <- ask_for(corp, "Phonetic =~ [aeiou]")

# 3. Extract formant measurements
formants <- quantify(vowels, superassp::forest)

# 4. Add speaker metadata
formants_with_meta <- biographize(formants, corp)

# 5. Analyze
library(dplyr)
formants_with_meta %>%
  group_by(label, Gender) %>%
  summarise(
    mean_F1 = mean(F1, na.rm = TRUE),
    mean_F2 = mean(F2, na.rm = TRUE)
  )
```

That's it! Let's break down each step.

## Step 1: Loading Your Corpus

The `corpus()` function is your entry point to working with EMU databases in reindeer.

### Basic Usage

```{r}
# From a path
corp <- corpus("path/to/database_emuDB")

# With verbose output to see what's happening
corp <- corpus("path/to/database_emuDB", verbose = TRUE)
```

### What Happens When You Load

When you create a corpus object, reindeer:

1. **Validates** the database structure
2. **Builds a SQLite cache** for fast queries
3. **Loads metadata** from .meta_json files
4. **Initializes** persistent connections

```{r}
# The corpus object contains
corp@dbName        # Database name
corp@basePath      # Full path to database
corp@config        # Database configuration
```

### Working with emuR

reindeer is fully compatible with emuR:

```{r}
# You can also load with emuR first
handle <- emuR::load_emuDB("path/to/database_emuDB")

# Then convert to reindeer corpus
corp <- corpus(handle)
```

## Step 2: Querying Your Data

reindeer provides optimized querying through `ask_for()` (alias: `query()`).

### Simple Queries

```{r}
# Exact match
segments <- ask_for(corp, "Phonetic == t")

# Not equal
segments <- ask_for(corp, "Phonetic != t")

# Regex pattern matching
vowels <- ask_for(corp, "Phonetic =~ [aeiou]")
consonants <- ask_for(corp, "Phonetic !~ [aeiou]")
```

### Advanced Queries

```{r}
# Sequence: find /t/ followed by vowel
seq <- ask_for(corp, "[Phonetic == t -> Phonetic =~ [aeiou]]")

# Dominance: find vowels dominated by stressed syllables
stressed_vowels <- ask_for(corp, "[Syllable == S ^ Phonetic =~ [aeiou]]")

# Conjunction (AND)
long_vowels <- ask_for(corp, "[Phonetic =~ [aeiou] & Duration > 100]")

# Disjunction (OR)
stops <- ask_for(corp, "[Phonetic == p | Phonetic == t | Phonetic == k]")
```

### Projection

Use `#` to specify which level to return:

```{r}
# Return the syllable, not the vowel
syllables <- ask_for(corp, "[Syllable #== S ^ Phonetic =~ [aeiou]]")
```

### Query Results

The result is a `segment_list` object:

```{r}
segments <- ask_for(corp, "Phonetic == t")

# View results
print(segments)      # Pretty printed summary
summary(segments)    # Detailed statistics

# Access as data frame
as.data.frame(segments)

# Key columns
segments$labels      # Segment labels
segments$start       # Start time (ms)
segments$end         # End time (ms)
segments$session     # Session name
segments$bundle      # Bundle name
```

## Step 3: Signal Processing (Quantify)

Extract acoustic measurements from your segments with `quantify()`.

### Basic Formant Analysis

```{r}
# Extract formants at segment midpoint
formants <- quantify(segments, superassp::forest)

# Result is an extended_segment_list
formants$F1          # First formant
formants$F2          # Second formant
formants$F3          # Third formant
```

### Multiple Time Points

```{r}
# Extract at 20%, 50%, and 80% of segment duration
formants_multi <- quantify(
  segments,
  superassp::forest,
  .at = c(0.2, 0.5, 0.8)
)

# Now you have 3 rows per segment
nrow(segments)           # e.g., 100
nrow(formants_multi)     # e.g., 300 (3 per segment)
```

### Different Analyses

```{r}
# Pitch (F0)
pitch <- quantify(segments, superassp::ksvF0)

# Intensity
intensity <- quantify(segments, superassp::rmsana)

# Zero crossing rate
zcr <- quantify(segments, superassp::zcrana)

# Spectrum
spectrum <- quantify(segments, superassp::dftSpectrum)
```

### Custom Parameters

```{r}
# Override default parameters
formants_custom <- quantify(
  segments,
  superassp::forest,
  nominalF1 = 500,     # Expected F1 (Hz)
  windowSize = 20,     # Window size (ms)
  numFormants = 4      # Extract 4 formants
)
```

### Metadata-Driven Parameters

reindeer automatically adjusts parameters based on speaker metadata:

```{r}
# If you've set Age and Gender metadata, these affect:
# - nominalF1, nominalF2, nominalF3 (formants)
# - Window sizes
# - Pitch ranges

# See default parameters
data(DSPP)
head(DSPP)
```

### Performance Options

```{r}
# Use caching for expensive analyses
formants <- quantify(
  segments,
  superassp::forest,
  .use_cache = TRUE,
  .cache_format = "qs"  # Fast serialization
)

# Parallel processing (default for large segment lists)
formants <- quantify(
  segments,
  superassp::forest,
  .parallel = TRUE,
  .workers = 4
)

# Verbose output
formants <- quantify(
  segments,
  superassp::forest,
  .verbose = TRUE
)
```

## Step 4: Adding Metadata

The `biographize()` function enriches your data with speaker metadata.

### Basic Usage

```{r}
# Add all available metadata
data_with_meta <- biographize(formants, corp)

# Now you have speaker information
data_with_meta$Age
data_with_meta$Gender
data_with_meta$Project
# ... any metadata you've defined
```

### Setting Metadata

Before you can use metadata, you need to set it:

```{r}
# Database-level (applies to all)
add_metadata(corp, list(
  Project = "MyStudy",
  SampleRate = 44100
))

# Session-level (overrides database defaults)
add_metadata(corp,
  list(Speaker = "P01", Age = 25, Gender = "M"),
  session = "Session1"
)

# Bundle-level (overrides session defaults)
add_metadata(corp,
  list(Quality = "Good", Notes = "Clear recording"),
  session = "Session1",
  bundle = "Bundle1"
)
```

### Metadata Inheritance

Reindeer uses a three-level hierarchy:

1. **Database** → defaults for everything
2. **Session** → overrides database
3. **Bundle** → overrides session

```{r}
# View effective metadata for a bundle
get_metadata(corp, session = "Session1", bundle = "Bundle1")
```

### Batch Metadata Management

```{r}
# Export to Excel for editing
export_metadata(corp, "metadata.xlsx")

# Edit in Excel...

# Import back
import_metadata(corp, "metadata.xlsx")

# Refresh cache
gather_metadata(corp)
```

## Step 5: Analysis and Visualization

Once you have your enriched data, use standard R tools:

```{r}
library(dplyr)
library(ggplot2)

# Summary statistics
formants_with_meta %>%
  group_by(label, Gender) %>%
  summarise(
    n = n(),
    mean_F1 = mean(F1, na.rm = TRUE),
    sd_F1 = sd(F1, na.rm = TRUE),
    mean_F2 = mean(F2, na.rm = TRUE),
    sd_F2 = sd(F2, na.rm = TRUE)
  )

# Vowel plot
ggplot(formants_with_meta, aes(x = F2, y = F1, color = label)) +
  geom_point(alpha = 0.5) +
  scale_x_reverse() +
  scale_y_reverse() +
  facet_wrap(~ Gender) +
  theme_minimal() +
  labs(title = "Vowel Space by Gender")

# Distribution plots
ggplot(formants_with_meta, aes(x = F1, fill = Gender)) +
  geom_density(alpha = 0.5) +
  facet_wrap(~ label) +
  theme_minimal()
```

## Common Workflows

### Vowel Analysis

```{r}
# Complete vowel analysis workflow
corp <- corpus("path/to/db_emuDB")

# Query vowels
vowels <- ask_for(corp, "Phonetic =~ [aeiouæɑɔ]")

# Extract formants at midpoint
formants <- quantify(vowels, superassp::forest, .at = 0.5)

# Add metadata
vowels_data <- biographize(formants, corp)

# Normalize formants
library(vowels)
vowels_data$F1_norm <- norm.lobanov(vowels_data$F1,
                                     vowels_data$Speaker)
vowels_data$F2_norm <- norm.lobanov(vowels_data$F2,
                                     vowels_data$Speaker)

# Plot
ggplot(vowels_data, aes(x = F2_norm, y = F1_norm, color = label)) +
  geom_point(alpha = 0.6) +
  stat_ellipse(level = 0.67) +
  scale_x_reverse() +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "Normalized Vowel Space")
```

### Pitch Analysis

```{r}
# Analyze pitch contours
corp <- corpus("path/to/db_emuDB")

# Query intonation phrases
phrases <- ask_for(corp, "Intonation == IP")

# Extract pitch at multiple points
pitch_contour <- quantify(
  phrases,
  superassp::ksvF0,
  .at = seq(0, 1, by = 0.1)  # Every 10%
)

# Add metadata
pitch_data <- biographize(pitch_contour, corp)

# Visualize
ggplot(pitch_data, aes(x = time_rel, y = F0, group = bundle)) +
  geom_line(alpha = 0.3) +
  stat_summary(fun = mean, geom = "line", color = "red", size = 1) +
  facet_wrap(~ Gender) +
  theme_minimal() +
  labs(
    title = "Pitch Contours in Intonation Phrases",
    x = "Relative Time",
    y = "F0 (Hz)"
  )
```

### VOT Measurement

```{r}
# Voice Onset Time for stops
corp <- corpus("path/to/db_emuDB")

# Query stops at word onsets
stops <- ask_for(corp, "[Word #^ Phonetic =~ [ptk]]")

# Get zero crossing rate (for burst detection)
zcr <- quantify(
  stops,
  superassp::zcrana,
  .at = seq(0, 0.2, by = 0.01)  # First 20% in 1% steps
)

# Add metadata
vot_data <- biographize(zcr, corp)

# Detect burst (simplified)
library(tidyr)
vot_summary <- vot_data %>%
  group_by(session, bundle, label) %>%
  mutate(
    burst_point = which.max(zcr)[1],
    vot_estimate = time_rel[burst_point] * (end - start)
  ) %>%
  slice(1) %>%
  ungroup()

# Compare by place of articulation
ggplot(vot_summary, aes(x = label, y = vot_estimate)) +
  geom_boxplot() +
  theme_minimal() +
  labs(
    title = "VOT by Place of Articulation",
    x = "Phoneme",
    y = "VOT (ms)"
  )
```

## Tips and Best Practices

### 1. Use Caching

For large corpora or expensive analyses:

```{r}
# Enable caching
formants <- quantify(segments, superassp::forest, .use_cache = TRUE)

# Check cache status
manage_cache(corp)

# Clean old caches
manage_cache(corp, action = "clean", days_old = 30)
```

### 2. Leverage Metadata

Set up metadata early:

```{r}
# Export template
export_metadata(corp, "metadata_template.xlsx")

# Fill in Excel, then import
import_metadata(corp, "metadata_filled.xlsx")
```

### 3. Use Verbose Mode for Debugging

```{r}
corp <- corpus("path/to/db_emuDB", verbose = TRUE)
formants <- quantify(segments, superassp::forest, .verbose = TRUE)
```

### 4. Check Data Before Analysis

```{r}
# Summary statistics
summary(segments)

# Check for missing data
sum(is.na(formants$F1))

# Visualize distributions
hist(formants$F1)
```

### 5. Optimize Queries

```{r}
# Be specific in queries to reduce processing time
vowels <- ask_for(corp, "Phonetic =~ [aeiou]")  # Good
all_segs <- ask_for(corp, "Phonetic =~ .*")     # Slower
```

## Troubleshooting

### Common Issues

**Problem:** `corpus()` fails with "Database path does not exist"

```{r}
# Solution: Check path
dir.exists("path/to/database_emuDB")

# Make sure it ends with _emuDB
corp <- corpus("path/to/database_emuDB")
```

**Problem:** `ask_for()` returns empty segment_list

```{r}
# Check your query syntax
# Common issues:
ask_for(corp, "Phonetic = t")    # Wrong: use ==
ask_for(corp, "Phonetic == t")   # Correct

# Check what levels exist
corp@config$levelDefinitions
```

**Problem:** `quantify()` is very slow

```{r}
# Enable caching
formants <- quantify(segments, superassp::forest, .use_cache = TRUE)

# Use parallel processing
formants <- quantify(segments, superassp::forest, .parallel = TRUE)

# Reduce number of time points
formants <- quantify(segments, superassp::forest, .at = 0.5)  # Just midpoint
```

**Problem:** Missing metadata after `biographize()`

```{r}
# Check if metadata exists
get_metadata(corp)

# Refresh metadata cache
gather_metadata(corp)

# Add metadata if missing
add_metadata(corp, list(Gender = "M"), session = "Session1")
```

## Next Steps

Now that you know the basics, explore:

- **[metadata_management.Rmd](metadata_management.html)** - Deep dive into metadata
- **[cache_management.Rmd](cache_management.html)** - Optimize performance
- **[simulation-infrastructure.Rmd](simulation-infrastructure.html)** - Parameter exploration
- **[query_benchmarks.qmd](query_benchmarks.html)** - Query optimization

## Getting Help

```{r}
# Function documentation
?corpus
?ask_for
?quantify
?biographize

# Package overview
help(package = "reindeer")

# Vignettes
browseVignettes("reindeer")
```

## Summary

The reindeer workflow in 4 steps:

1. **Load:** `corp <- corpus("path/to/db_emuDB")`
2. **Query:** `segments <- ask_for(corp, "Phonetic == t")`
3. **Analyze:** `formants <- quantify(segments, superassp::forest)`
4. **Enrich:** `data <- biographize(formants, corp)`

Happy analyzing! 🦌
