---
title: "Getting Started with reindeer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with reindeer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# The measurement chunks need a signal-processing companion; the rest runs on
# the corpus bundled with the package, so this vignette works offline.
eval_dsp <- requireNamespace("superassp", quietly = TRUE)
```

## What reindeer does for you

**reindeer** lets you treat an EMU-SDMS speech corpus the way you treat
any tidy dataset:

- Open a corpus with `corpus()`.
- Pull out annotation segments with `query()` using EMU Query Language.
- Add measurements (formants, pitch, intensity…) with `quantify()`,
  picking parameters from each speaker's metadata automatically.
- Join speaker metadata with `biographize(segs, corp)`.
- Then plot, model, or write to disk with the dplyr verbs you already
  know.

This vignette shows the five-minute version of that loop, then expands
each step. Everything below runs against the demo corpus that ships with
the package. The measurement steps additionally need **superassp**; where
it is not installed those chunks are shown but not executed.

```{r install, eval = FALSE}
# Installation
remotes::install_github("humlab-speech/reindeer")
library(reindeer)
```

## Five-minute workflow

```{r workflow}
library(reindeer)
library(dplyr)

corp <- demo_corpus()

# 1. Query vowels
vowels <- query(corp, "Phonetic =~ [aeiou]")
```

```{r workflow-quantify, eval = eval_dsp}
# 2. Extract formants at the midpoint
formants <- quantify(vowels, superassp::trk_formant_forest, .at = 0.5)

# 3. Add speaker metadata
data <- biographize(formants, corp)

# 4. Summarise
reindeer::collect(data) |>
  group_by(labels) |>
  summarise(mean_F1_Hz = mean(F1_Hz, na.rm = TRUE),
            mean_F2_Hz = mean(F2_Hz, na.rm = TRUE),
            .groups = "drop")
```

That's the whole pipeline. The rest of the vignette walks through each
verb in more detail.

## Step 1 — Open the corpus

```{r open, eval = FALSE}
corp <- corpus("path/to/your_emuDB")   # your corpus goes here
```

This vignette uses the demo corpus instead, so it runs anywhere:

```{r open-demo}
corp <- demo_corpus()
corp
```

The trailing `_emuDB` is appended automatically if you leave it off. If
the corpus does not exist yet, pass `create = TRUE`.

## Step 2 — Query

`query()` understands EMU Query Language (EQL). Common shapes:

```{r query-shapes, eval = FALSE}
query(corp, "Phonetic == t")                        # exact match
query(corp, "Phonetic =~ [aeiou]")                  # regex
query(corp, "[Phonetic == t -> Phonetic =~ [aeiou]]")     # sequence
query(corp, "[Syllable #== S ^ Phonetic =~ [aeiou]]")     # dominance
```

Use `==` for equality (`=` triggers a parse error), `=~` / `!~` for
regex, and wrap sequences and dominances in `[]`. See the
[EQL grammar](https://ips-lmu.github.io/The-EMU-SDMS-Manual/chap-querysys.html)
for the full reference.

Results are tidy: each row is one segment with `labels`, `start`, `end`,
`session`, `bundle`, plus a few bookkeeping columns. You can pipe the
result through any dplyr verb.

```{r query-pipe}
vowels <- query(corp, "Phonetic =~ [aeiou]", lazy = FALSE)
vowels |> filter(end - start > 100) |> count(labels)
```

## Step 3 — Quantify

`quantify()` runs a DSP function on each segment and returns one row of
measurements per segment (or per time point if you pass `.at`).

```{r quantify, eval = eval_dsp}
formants   <- quantify(vowels, superassp::trk_formant_forest, .at = 0.5)
contour    <- quantify(vowels, superassp::trk_pitch_ksv, .at = seq(0, 1, 0.1))
intensity  <- quantify(vowels, superassp::trk_rms)
head(formants)
```

By default reindeer chooses age- and gender-aware parameters from each
speaker's metadata (think: smaller windows for children, lower expected
F1 for adult males). Preview what is being chosen with
`dsp_parameters()`:

```{r dsp-params, eval = eval_dsp}
dsp_parameters(age = 8,  gender = "Male")
dsp_parameters(age = 45, gender = "Female")
```

Override anything explicitly through `...`:

```{r quantify-override, eval = FALSE}
quantify(vowels, superassp::trk_formant_forest,
         nominalF1 = 500, windowSize = 20, numFormants = 4)
```

### Caching results

Slow DSP runs are worth caching. Turn it on once and reindeer skips work
that is already on disk:

```{r cache, eval = eval_dsp}
formants <- quantify(vowels, superassp::trk_formant_forest,
                     .at = 0.5, .use_cache = TRUE)
table(formants$.cache_status)   # "hit" / "miss" per row
inspect_cache(corp)             # see what's stored
```

## Step 4 — Add speaker metadata

`biographize()` joins corpus metadata onto a segment_list; `quantify()`
does the DSP extraction. They're two separate verbs — no mode switch needed.

```{r enrich, eval = eval_dsp}
data <- biographize(formants, corp)         # metadata join
names(data)
```

To set metadata in the first place, use `set_metadata()`. The three
levels form an inheritance chain:

```{r metadata}
set_metadata(corp, list(Project = "MyStudy"))                  # database
set_metadata(corp, list(Speaker = "P001", Age = 25,
                        Gender = "Female"))                    # database-wide
get_metadata(corp) |> select(session, bundle, Project, Speaker, Age, Gender)
```

Bundle wins over session, session wins over database default.
`load_metadata(corp)` rescans the on-disk `METADATA.json` files into the
cache when you have edited them outside R. For batch editing,
round-trip through Excel:

```{r metadata-excel, eval = FALSE}
export_metadata(corp, "metadata.xlsx")        # edit in spreadsheet
load_metadata(corp, source = "excel", path = "metadata.xlsx")
```

## Step 5 — Analyse

Once you have a tibble with measurements and metadata, the rest is
ordinary R.

```{r plot, fig.width = 6, fig.height = 4, eval = eval_dsp}
library(ggplot2)

reindeer::collect(data) |>
  ggplot(aes(x = as.numeric(F2_Hz), y = as.numeric(F1_Hz), color = Gender)) +
  geom_point(alpha = 0.5) +
  scale_x_reverse() +
  scale_y_reverse() +
  theme_minimal() +
  labs(title = "Vowel space")
```

## Interactive annotation (optional)

To inspect or edit annotations in a browser-based UI:

```{r serve, eval = FALSE}
serve_app(corp)                              # whole corpus
serve_app(corp, seglist = vowels)            # restrict to a query result
serve_app(corp, sessionPattern = "Session0[1-3]")
```

`serve_app()` starts a local EMU-webApp instance and returns; close the
tab and call `httpuv::stopAllServers()` to release the port. It is not
evaluated here because it opens a browser and a socket.

## Lazy pipelines and pipe-loss

`query()` is lazy by default — it returns a `lazy_segment_list` and
nothing actually runs until you call `nrow()`, `head()`, `collect()`,
or any data-access method. Filter and navigate freely:

```{r lazy}
plan <- query(corp, "Phonetic =~ [aeiou]") |>
  filter(labels != "@") |>
  scout(steps_forward = 1) |>      # next phoneme
  ascend_to("Syllable")            # containing syllable

result <- reindeer::collect(plan)
nrow(result)
```

If a pipeline drops more than 25% of rows reindeer warns. Inspect the
loss with:

```{r provenance}
provenance(plan)        # step / verb / rows_in / rows_out / rows_lost
dropped_rows(plan)      # per-step row-loss summary
dropped(plan)           # cumulative count
```

## Where to go next

- `vignette("metadata_management")` — three-level inheritance,
  spreadsheet round-trip, validation.
- `vignette("cache_management")` — inspect_cache, cleanup, thresholds.
- `vignette("interactive_annotation")` — full serve / EMU-webApp workflow.
- `vignette("lazy_and_provenance")` — lazy plans, pipe-loss debugging.
- `?corpus`, `?query`, `?quantify`, `?enrich`, `?set_metadata` —
  function reference.
