---
title: "Getting Started with reindeer"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with reindeer}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = identical(Sys.getenv("REINDEER_EVAL_VIGNETTES"), "true")
)
```

## What reindeer does for you

**reindeer** lets you treat an EMU-SDMS speech corpus the way you treat
any tidy dataset:

- Open a corpus with `corpus()`.
- Pull out annotation segments with `query()` using EMU Query Language.
- Add measurements (formants, pitch, intensity…) with `quantify()`,
  picking parameters from each speaker's metadata automatically.
- Join speaker metadata with `enrich(..., with = "metadata")`.
- Then plot, model, or write to disk with the dplyr verbs you already
  know.

This vignette shows the five-minute version of that loop, then expands
each step.

## Installation

```{r}
remotes::install_github("humlab-speech/reindeer")
library(reindeer)
```

## Five-minute workflow

```{r}
library(reindeer)
library(dplyr)

corp <- corpus("path/to/your_emuDB")

# 1. Query vowels
vowels <- query(corp, "Phonetic =~ [aeiou]")

# 2. Extract formants at the midpoint
formants <- quantify(vowels, superassp::forest, .at = 0.5)

# 3. Add speaker metadata
data <- enrich(formants, corp)

# 4. Summarise
data |>
  group_by(label, Gender) |>
  summarise(mean_F1 = mean(F1, na.rm = TRUE),
            mean_F2 = mean(F2, na.rm = TRUE),
            .groups = "drop")
```

That's the whole pipeline. The rest of the vignette walks through each
verb in more detail.

## Step 1 — Open the corpus

```{r}
corp <- corpus("path/to/your_emuDB")
```

The trailing `_emuDB` is appended automatically if you leave it off.
If the corpus does not exist yet, pass `create = TRUE`.

## Step 2 — Query

`query()` understands EMU Query Language (EQL). Common shapes:

```{r}
query(corp, "Phonetic == t")                       # exact match
query(corp, "Phonetic =~ [aeiou]")                 # regex
query(corp, "[Phonetic == t -> Phonetic =~ [aeiou]]")    # sequence
query(corp, "[Syllable #== S ^ Phonetic =~ [aeiou]]")    # dominance
```

Use `==` for equality (`=` triggers a parse error), `=~` / `!~` for
regex, and wrap sequences and dominances in `[ ]`. See the
[EQL grammar](https://ips-lmu.github.io/The-EMU-SDMS-Manual/chap-querysys.html)
for the full reference.

Results are tidy: each row is one segment with `label`, `start`, `end`,
`session`, `bundle`, plus a few bookkeeping columns. You can pipe the
result through any dplyr verb.

```{r}
vowels <- query(corp, "Phonetic =~ [aeiou]")
vowels |> filter(end - start > 100) |> count(label)
```

## Step 3 — Quantify

`quantify()` runs a DSP function on each segment and returns one row of
measurements per segment (or per time point if you pass `.at`).

```{r}
formants <- quantify(vowels, superassp::forest)                 # midpoint
contour  <- quantify(vowels, superassp::ksvF0, .at = seq(0, 1, 0.1))
intensity <- quantify(vowels, superassp::rmsana)
```

By default reindeer chooses age- and gender-aware parameters from each
speaker's metadata (think: smaller windows for children, lower expected
F1 for adult males). Preview what's being chosen with
`dsp_parameters()`:

```{r}
dsp_parameters(age = 8, gender = "Male")
dsp_parameters(corpus_obj = corp)   # one row per bundle
```

Override anything explicitly through `...`:

```{r}
quantify(vowels, superassp::forest,
         nominalF1 = 500, windowSize = 20, numFormants = 4)
```

### Caching results

Slow DSP runs are worth caching. Turn it on once and reindeer will skip
work that's already on disk:

```{r}
formants <- quantify(vowels, superassp::forest, .use_cache = TRUE)
formants$.cache_status   # "hit" / "miss" per row

inspect_cache(corp)      # see what's stored
```

## Step 4 — Add speaker metadata

`enrich()` does double duty: with `with = "metadata"` it joins corpus
metadata onto the segments; with `.using = fn` it forwards to
`quantify()`.

```{r}
data <- enrich(formants, corp)              # metadata join (default)
```

To set metadata in the first place, use `set_metadata()`:

```{r}
set_metadata(corp, list(Project = "MyStudy"))                       # db level
set_metadata(corp, list(Age = 25, Gender = "Female"),
             session = "Session1")                                  # session
set_metadata(corp, list(Quality = "Good"),
             session = "Session1", bundle = "Bundle1")              # bundle
```

Bundle wins over session, session wins over database default.
`load_metadata(corp)` rescans the on-disk `METADATA.json` files into
the cache when you have edited them outside R. For batch editing,
round-trip through Excel:

```{r}
export_metadata(corp, "metadata.xlsx")        # edit in spreadsheet
load_metadata(corp, source = "excel", path = "metadata.xlsx")
```

## Step 5 — Analyse

Once you have a tibble with measurements and metadata, the rest is
ordinary R.

```{r}
library(ggplot2)

ggplot(data, aes(x = F2, y = F1, color = label)) +
  geom_point(alpha = 0.5) +
  scale_x_reverse() +
  scale_y_reverse() +
  facet_wrap(~ Gender) +
  theme_minimal() +
  labs(title = "Vowel space by gender")
```

## Interactive annotation (optional)

To inspect or edit annotations in a browser-based UI:

```{r}
serve_app(corp)                              # whole corpus
serve_app(corp, seglist = vowels)            # restrict to query result
serve_app(corp, sessionPattern = "Session0[1-3]")
```

Close the tab or run `httpuv::stopAllServers()` to release the port.

## Lazy pipelines and pipe-loss

`query()` is lazy by default — it returns a `lazy_segment_list` and
nothing actually runs until you call `nrow()`, `head()`, `collect()`,
or any data-access method. Filter and navigate freely:

```{r}
plan <- query(corp, "Phonetic =~ [aeiou]") |>
  filter(label != "@") |>
  scout(steps_forward = 1) |>      # next phoneme
  ascend_to("Word")                # parent word

collect(plan)                       # run it
```

If a pipeline drops more than 25% of rows reindeer warns. Inspect the
loss with:

```{r}
provenance(plan)        # tibble: step / verb / rows_in / rows_out / rows_lost
dropped_rows(plan)      # per-step row-loss summary
dropped(plan)           # cumulative count
```

## Where to go next

- `vignette("metadata_management")` — three-level inheritance,
  spreadsheet round-trip, validation.
- `vignette("cache_management")` — inspect_cache, cleanup, thresholds.
- `vignette("interactive_annotation")` — full serve / EMU-webApp workflow.
- `vignette("lazy_and_provenance")` — lazy plans, pipe-loss debugging.
- `?corpus`, `?query`, `?quantify`, `?enrich`, `?set_metadata` —
  function reference.
