---
title: "Interactive Annotation with EMU-webApp"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Annotation with EMU-webApp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

The **reindeer** package provides `serve()`, a powerful function that launches the EMU-webApp annotation interface directly from R. This allows you to:

- **Visualize** your speech data with interactive spectrograms
- **Annotate** segments, events, and hierarchical structures
- **Play audio** synchronized with annotations
- **Edit** annotations in real-time
- **Save** changes directly back to your database

This vignette shows you how to use `serve()` for interactive annotation workflows.

## Quick Start

```{r quick-start}
library(reindeer)

# Load your corpus
corp <- corpus("path/to/database_emuDB")

# Launch the annotation interface
serve(corp)
```

That's it! The EMU-webApp will open in your browser (or RStudio Viewer pane) and you can start annotating.

## Setup and Configuration

### First-Time Setup

The `serve()` function looks for the EMU-webApp in several locations:

```{r setup}
# Method 1: Set R option (recommended for interactive use)
options(reindeer.emuWebApp.dir = "/path/to/EMU-webApp/dist")

# Method 2: Set environment variable (recommended for scripts)
Sys.setenv(EMU_WEBAPP_DIR = "/path/to/EMU-webApp/dist")

# Method 3: Let reindeer find it automatically
# - Checks package installation directory
# - Looks for ../EMU-webApp/dist relative to package
serve(corp)  # Will use auto-detected path
```

### Verifying Setup

```{r verify}
# The function will show an informative error if webapp not found
serve(corp)
# Error: EMU-webApp directory not found.
# ℹ Tried: /path/that/was/checked
# ℹ Set a custom path via options(reindeer.emuWebApp.dir = '/path/to/EMU-webApp/dist')
```

## Basic Annotation Workflows

### Annotate Entire Corpus

```{r annotate-all}
corp <- corpus("path/to/database_emuDB")

# Serve all bundles
serve(corp)

# The webApp will show:
# - All sessions in the corpus
# - All bundles in each session
# - Navigate with arrow keys or bundle list
```

### Annotate Specific Sessions

```{r annotate-sessions}
# Serve only specific sessions using regex patterns
serve(corp, sessionPattern = "Session0[12]")

# Or more complex patterns
serve(corp, sessionPattern = "^(pilot|main)_")
```

### Annotate Specific Bundles

```{r annotate-bundles}
# Serve only bundles matching a pattern
serve(corp, bundlePattern = "recording_0[1-5]")

# Combine session and bundle patterns
serve(corp,
      sessionPattern = "Session01",
      bundlePattern = "msajc.*")
```

### Annotate Query Results

This is particularly powerful - navigate directly to segments of interest:

```{r annotate-query}
# Find all /t/ phonemes
t_segments <- query(corp, "Phonetic == t")

# Open webApp at those specific segments
serve(corp, seglist = t_segments)

# The webApp will:
# 1. Show only bundles containing /t/
# 2. Automatically navigate to each /t/ segment
# 3. Highlight the segment boundaries
```

### Workflow Example: Annotate Vowels

```{r vowel-workflow}
# 1. Query for vowels
vowels <- query(corp, "Phonetic =~ [aeiou]")

# 2. Check how many bundles
length(unique(paste(vowels$session, vowels$bundle)))
# > 45 bundles with vowels

# 3. Open annotation interface
serve(corp, seglist = vowels)

# 4. Annotate in the webApp:
#    - Add stress markers
#    - Adjust boundaries
#    - Add comments

# 5. Changes are saved automatically to the database
```

## Advanced Usage

### Custom Server Settings

```{r custom-server}
# Use a different port (useful if default is busy)
serve(corp, port = 8080)

# Bind to a different interface
serve(corp, host = "0.0.0.0", port = 17890)  # Allow external connections

# Disable auto-opening browser
serve(corp, autoOpenURL = "")

# Use external browser instead of RStudio viewer
serve(corp, useViewer = FALSE)
```

### Debug Mode

```{r debug}
# Enable debug output
serve(corp, debug = TRUE, debugLevel = 2)

# Debug levels:
# 0 = no debug output
# 1 = basic info
# 2 = detailed protocol messages
# 8 = maximum verbosity
```

### Working with Bundle Lists

If your database has predefined bundle lists:

```{r bundle-lists}
# Serve a specific bundle list
serve(corp, bundleListName = "AnnotationQueue")

# This is useful for:
# - Annotation workflows with specific bundles to review
# - Quality control checklists
# - Collaborative annotation (different annotators get different lists)
```

## Annotation Interface Features

### Navigation

Once the webApp opens:

**Keyboard Shortcuts:**
- `←` / `→` - Previous/Next bundle
- `Space` - Play/Pause audio
- `Ctrl + ←` / `Ctrl + →` - Move 10 seconds
- `Ctrl + Shift + ←` / `Ctrl + Shift + →` - Move to start/end

**Mouse:**
- Click waveform - Set playback position
- Drag on spectrogram - Create new segment
- Click segment boundary - Adjust boundary
- Right-click segment - Edit label or delete

### Editing Annotations

**Add Segments:**
1. Select the target level in the level panel
2. Drag on the spectrogram to create boundaries
3. Type label in the popup
4. Press Enter to confirm

**Modify Segments:**
1. Click segment boundary to drag it
2. Double-click label to edit text
3. Right-click → Delete to remove

**Hierarchical Annotation:**
- The interface shows parent-child relationships
- Drag to link items across levels
- Edit link types in the context menu

### Signal Tracks

View multiple signal representations:

**Default Tracks:**
- Waveform (OSCI)
- Spectrogram (SPEC)

**Add Custom Tracks:**
```{r custom-tracks}
# These must be pre-computed and added to the database
# via emuR::add_ssffTrackDefinition()

# Example: Add formant track
# Then it will appear in the webApp automatically
```

## Stopping the Server

**Method 1: Clear button in webApp**
- Click the "clear" button in the EMU-webApp interface

**Method 2: Close browser**
- Simply close or reload the webApp page

**Method 3: R command**
```{r stop-server}
# Stop all servers
httpuv::stopAllServers()

# Or restart R session
.rs.restartR()  # In RStudio
```

## Quality Control Workflow

Here's a complete workflow for annotation QC:

```{r qc-workflow}
library(reindeer)
library(dplyr)

# 1. Load corpus
corp <- corpus("path/to/database_emuDB")

# 2. Find segments that might need review
#    (e.g., very short segments)
all_segments <- query(corp, "Phonetic =~ .*")

short_segments <- all_segments %>%
  mutate(duration = end - start) %>%
  filter(duration < 20)  # Less than 20ms

cat(sprintf("Found %d short segments to review\n", nrow(short_segments)))

# 3. Open annotation interface for review
serve(corp, seglist = short_segments)

# 4. Annotator reviews and fixes in webApp
#    - Adjust boundaries
#    - Delete artifacts
#    - Merge segments

# 5. Re-query to verify fixes
all_segments_updated <- query(corp, "Phonetic =~ .*")

short_segments_updated <- all_segments_updated %>%
  mutate(duration = end - start) %>%
  filter(duration < 20)

cat(sprintf("Now %d short segments remain\n", nrow(short_segments_updated)))
```

## Integration with Analysis Pipeline

The `serve()` function integrates seamlessly with analysis:

```{r integration}
# Analysis → Annotation → Re-analysis cycle

# 1. Initial analysis
corp <- corpus("path/to/database_emuDB")
vowels <- query(corp, "Phonetic =~ [aeiou]")
formants <- quantify(vowels, superassp::forest)

# 2. Find outliers
library(dplyr)
outliers <- formants %>%
  group_by(label) %>%
  mutate(
    F1_z = scale(F1),
    F2_z = scale(F2)
  ) %>%
  filter(abs(F1_z) > 3 | abs(F2_z) > 3) %>%
  ungroup()

cat(sprintf("Found %d outlier measurements\n", nrow(outliers)))

# 3. Review outliers in context
serve(corp, seglist = outliers)

# Annotator checks:
# - Are boundaries accurate?
# - Is the label correct?
# - Is there background noise?
# - Should this token be excluded?

# 4. Re-run analysis after corrections
vowels_corrected <- query(corp, "Phonetic =~ [aeiou]")
formants_corrected <- quantify(vowels_corrected, superassp::forest)

# 5. Compare results
cat("Before correction:\n")
summary(formants$F1)

cat("\nAfter correction:\n")
summary(formants_corrected$F1)
```

## Collaborative Annotation

For team annotation projects:

```{r collaborative}
# 1. Set up bundle lists for each annotator
corp <- corpus("path/to/database_emuDB")

# Create bundle lists (one-time setup, using emuR)
# emuR::add_bundleList(corp, "Annotator_A", bundles_A)
# emuR::add_bundleList(corp, "Annotator_B", bundles_B)

# 2. Each annotator serves their assigned list
serve(corp, bundleListName = "Annotator_A")

# 3. Track progress
bundles_A <- emuR::read_bundleList(get_emuDBhandle(corp), "Annotator_A")

# Check which have been annotated
# (Look for .meta_json or specific annotation levels)
```

## Troubleshooting

### Common Issues

**Problem:** WebApp doesn't open

```{r troubleshoot-open}
# Check if port is already in use
serve(corp, port = 8080)  # Try different port

# Check browser setting
serve(corp, browser = getOption("browser"))
```

**Problem:** "EMU-webApp directory not found"

```{r troubleshoot-path}
# Set the path explicitly
options(reindeer.emuWebApp.dir = "/path/to/EMU-webApp/dist")

# Verify path exists
dir.exists(getOption("reindeer.emuWebApp.dir"))
```

**Problem:** Changes not saving

```{r troubleshoot-save}
# Ensure database isn't read-only
# Check file permissions on the database directory

# Verify changes are reflected
gather_metadata(corp, verbose = TRUE)
```

**Problem:** Server won't stop

```{r troubleshoot-stop}
# Force stop all servers
httpuv::stopAllServers()

# If that doesn't work, restart R session
.rs.restartR()  # RStudio only
```

### Getting Help

```{r help}
# Function documentation
?serve

# Check server status
# (No built-in status check, but server will error if already running)

# View EMU-webApp console for errors
# Open browser developer tools (F12)
```

## Tips and Best Practices

### 1. Start Small

```{r tip-small}
# Don't serve the entire corpus at once
# Start with a small session
serve(corp, sessionPattern = "pilot_session")

# Then scale up as needed
```

### 2. Use Query Results for Targeted Annotation

```{r tip-query}
# Much more efficient than browsing all bundles
vowels <- query(corp, "Phonetic =~ [aeiou]")
serve(corp, seglist = vowels)
```

### 3. Save Regularly

```{r tip-save}
# The webApp auto-saves, but:
# - Close properly (don't kill the server)
# - Check database after major annotation sessions
# - Consider git commits for version control
```

### 4. Use RStudio Viewer

```{r tip-viewer}
# Default behavior in RStudio
serve(corp, useViewer = TRUE)

# Benefits:
# - Side-by-side with R console
# - Easy to copy segment IDs or labels
# - Switch between code and annotation
```

### 5. Keyboard Shortcuts

Learn the keyboard shortcuts - they dramatically speed up annotation:
- Space: Play/pause
- Arrow keys: Navigate
- Enter: Confirm edit
- Esc: Cancel edit

### 6. Check Configuration

```{r tip-config}
# Before a long annotation session:
# 1. Verify correct levels are shown
corp@config$levelDefinitions

# 2. Check label definitions
corp@config$labelDefinitions

# 3. Ensure hierarchy is correct
corp@config$linkDefinitions
```

## Example: Complete Annotation Session

```{r complete-example}
library(reindeer)
library(dplyr)

# === SETUP ===
corp <- corpus("path/to/corpus_emuDB")

# === FIND WORK TO DO ===
# Find all words, check for missing phonetic annotations
words <- query(corp, "Word =~ .*")

# Check which bundles have words but no phonetic tier
bundles_with_words <- words %>%
  select(session, bundle) %>%
  distinct()

phonemes <- query(corp, "Phonetic =~ .*")
bundles_with_phonemes <- phonemes %>%
  select(session, bundle) %>%
  distinct()

bundles_needing_annotation <- anti_join(
  bundles_with_words,
  bundles_with_phonemes,
  by = c("session", "bundle")
)

cat(sprintf("Found %d bundles needing phonetic annotation\n",
            nrow(bundles_needing_annotation)))

# === ANNOTATE ===
# Convert to segment_list format for serve()
work_segments <- words %>%
  semi_join(bundles_needing_annotation, by = c("session", "bundle"))

# Open annotation interface
serve(corp, seglist = work_segments)

# [ Annotator works in webApp ]
# - Add phonetic segments within each word
# - Adjust boundaries as needed
# - Save changes (automatic)

# === VERIFY ===
# Check progress after annotation session
phonemes_updated <- query(corp, "Phonetic =~ .*")
bundles_with_phonemes_updated <- phonemes_updated %>%
  select(session, bundle) %>%
  distinct()

completed <- semi_join(
  bundles_needing_annotation,
  bundles_with_phonemes_updated,
  by = c("session", "bundle")
)

cat(sprintf("Completed %d/%d bundles\n",
            nrow(completed),
            nrow(bundles_needing_annotation)))

# === ANALYZE ===
# Now ready for acoustic analysis
phonemes_final <- query(corp, "Phonetic =~ [aeiou]")
formants <- quantify(phonemes_final, superassp::forest)

# Proceed with analysis...
```

## Summary

The `serve()` function provides:

- **Easy launch** - One function call to start annotating
- **Flexible filtering** - Serve exactly what you need to annotate
- **Query integration** - Jump directly to segments of interest
- **Seamless workflow** - Integrate annotation with analysis pipelines
- **Auto-save** - Changes persist to database automatically

This makes interactive annotation a natural part of your reindeer workflow, not a separate step with external tools.

## See Also

- `?serve` - Complete function documentation
- `vignette("getting_started")` - Reindeer basics
- `vignette("transcription_workflow")` - Automatic annotation
- EMU-webApp documentation: https://ips-lmu.github.io/EMU-webApp/
