---
title: "Interactive Annotation with Artic"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Interactive Annotation with Artic}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Every example below starts a local HTTP/WebSocket server and opens a
# browser, so none are evaluated here: a build must not launch servers or
# hijack the viewer. Chunks that can safely run opt back in with eval = TRUE.
knitr::opts_chunk$set(eval = FALSE)
```

## Introduction

The **reindeer** package provides `annotate()` and `review()`, which launch the [Artic](https://github.com/humlab-speech/artic) annotation interface directly from R. Artic shows interactive spectrograms and signal tracks, plays audio synchronised with the annotation, and writes edits straight back to the database. `annotate()` opens the whole corpus for editing; `review()` turns a segment list into a playlist of time anchors so you can step through just the portions you care about.

This vignette shows you how to use `annotate()` and `review()` for interactive annotation and review workflows.

## Quick start

```{r quick-start}
library(reindeer)

# Load your corpus
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")

# Launch the annotation interface
annotate(corp)
```

Artic opens in your browser (or the RStudio Viewer pane) and you can start annotating.

## What gets served

```{r quick-start-query, eval = TRUE}
# This chunk opts back in, so it must attach the package itself - the
# library() call above sits in a chunk that is deliberately not evaluated.
library(reindeer)
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")
vowels <- query(corp, "Phonetic =~ [aeiou]", lazy = FALSE)
dim(vowels)
head(vowels)[, c("session", "bundle", "labels", "start", "end")]
```

The corpus is narrowed by the arguments shown in the rest of this vignette -
`sessionPattern` and `bundlePattern` for `annotate()`, or a segment list from
`query()`/`quantify()` for `review()`.

## Setup and configuration

### First-time setup

`annotate()` and `review()` look for the Artic build in this order:

```{r setup}
# 1. Explicit per-call directory (wins over everything)
annotate(corp, appDir = "../artic/dist")

# 2. R option
options(reindeer.artic.dir = "/path/to/artic/dist")

# 3. Environment variable
Sys.setenv(ARTIC_DIR = "/path/to/artic/dist")

# 4. The build bundled with reindeer
# 5. The install_artic() cache
# 6. A sibling ../artic/dist checkout
annotate(corp)  # uses the auto-detected build
```

### Verifying setup

```{r verify}
# Inspect the resolved build and what its manifest advertises
artic_info()

# Install a different build: pinned download, or a local directory
install_artic()
install_artic(local = "../artic/dist")
```

## Basic annotation workflows

### Annotate entire corpus

```{r annotate-all}
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")

# Serve all bundles
annotate(corp)

# Artic will show:
# - All sessions in the corpus
# - All bundles in each session
# - Navigate with arrow keys or bundle list
```

### Annotate specific sessions

```{r annotate-sessions}
# Serve only specific sessions using regex patterns
annotate(corp, sessionPattern = "Session0[12]")

# Or more complex patterns
annotate(corp, sessionPattern = "^(pilot|main)_")
```

### Annotate specific bundles

```{r annotate-bundles}
# Serve only bundles matching a pattern
annotate(corp, bundlePattern = "recording_0[1-5]")

# Combine session and bundle patterns
annotate(corp,
      sessionPattern = "Session01",
      bundlePattern = "msajc.*")
```

### Annotate query results

You can navigate directly to the segments you care about:

```{r annotate-query}
# Find all /t/ phonemes
t_segments <- query(corp, "Phonetic == t")

# Open Artic at those specific segments
review(corp, t_segments)

# Artic will:
# 1. Show only bundles containing /t/
# 2. Automatically navigate to each /t/ segment
# 3. Highlight the segment boundaries
```

### Workflow example: annotate vowels

```{r vowel-workflow}
# 1. Query for vowels
vowels <- query(corp, "Phonetic =~ [aeiou]")

# 2. Check how many bundles
length(unique(paste(vowels$session, vowels$bundle)))
# > 45 bundles with vowels

# 3. Open annotation interface
review(corp, vowels)

# 4. Annotate in Artic:
#    - Add stress markers
#    - Adjust boundaries
#    - Add comments

# 5. Changes are saved automatically to the database
```

## Advanced usage

### Custom server settings

```{r custom-server}
# Use a different port (useful if default is busy)
annotate(corp, port = 8080)

# Bind to a different interface
annotate(corp, host = "0.0.0.0", port = 17890)  # Allow external connections

# Disable auto-opening browser
annotate(corp, autoOpenURL = "")

# Use external browser instead of RStudio viewer
annotate(corp, useViewer = FALSE)
```

### Debug mode

```{r debug}
# Enable debug output
annotate(corp, debug = TRUE, debugLevel = 2)

# Debug levels:
# 0 = no debug output
# 1 = basic info
# 2 = detailed protocol messages
# 8 = maximum verbosity
```

### Working with bundle lists

If your database has predefined bundle lists:

```{r bundle-lists}
# Serve a specific bundle list
annotate(corp, bundleListName = "AnnotationQueue")

# This is useful for:
# - Annotation workflows with specific bundles to review
# - Quality control checklists
# - Collaborative annotation (different annotators get different lists)
```

## Annotation interface features

### Navigation

Once Artic opens:

**Keyboard Shortcuts:**
- `←` / `→` - Previous/Next bundle
- `Space` - Play/Pause audio
- `Ctrl + ←` / `Ctrl + →` - Move 10 seconds
- `Ctrl + Shift + ←` / `Ctrl + Shift + →` - Move to start/end

**Mouse:**
- Click waveform - Set playback position
- Drag on spectrogram - Create new segment
- Click segment boundary - Adjust boundary
- Right-click segment - Edit label or delete

### Editing annotations

**Add Segments:**
1. Select the target level in the level panel
2. Drag on the spectrogram to create boundaries
3. Type label in the popup
4. Press Enter to confirm

**Modify Segments:**
1. Click segment boundary to drag it
2. Double-click label to edit text
3. Right-click → Delete to remove

**Hierarchical Annotation:**
- The interface shows parent-child relationships
- Drag to link items across levels
- Edit link types in the context menu

### Signal tracks

View multiple signal representations:

**Default Tracks:**
- Waveform (OSCI)
- Spectrogram (SPEC)

**Add Custom Tracks:**
```{r custom-tracks}
# These must be pre-computed and added to the database
# via emuR::add_ssffTrackDefinition()

# Example: Add formant track
# Then it will appear in Artic automatically
```

## Stopping the server

**Method 1: Clear button in Artic**
- Click the "clear" button in Artic interface

**Method 2: Close browser**
- Simply close or reload Artic page

**Method 3: R command**
```{r stop-server}
# Stop all servers
httpuv::stopAllServers()

# Or restart R session
.rs.restartR()  # In RStudio
```

## Quality control workflow

Here's a complete workflow for annotation QC:

```{r qc-workflow}
library(reindeer)
library(dplyr)

# 1. Load corpus
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")

# 2. Find segments that might need review
#    (e.g., very short segments)
all_segments <- query(corp, "Phonetic =~ .*")

short_segments <- all_segments %>%
  mutate(duration = end - start) %>%
  filter(duration < 20)  # Less than 20ms

cat(sprintf("Found %d short segments to review\n", nrow(short_segments)))

# 3. Open annotation interface for review
review(corp, short_segments)

# 4. Annotator reviews and fixes in Artic
#    - Adjust boundaries
#    - Delete artifacts
#    - Merge segments

# 5. Re-query to verify fixes
all_segments_updated <- query(corp, "Phonetic =~ .*")

short_segments_updated <- all_segments_updated %>%
  mutate(duration = end - start) %>%
  filter(duration < 20)

cat(sprintf("Now %d short segments remain\n", nrow(short_segments_updated)))
```

## Integration with analysis pipeline

`annotate()` integrates with an analysis pipeline:

```{r integration}
# Analysis → Annotation → Re-analysis cycle

# 1. Initial analysis
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")
vowels <- query(corp, "Phonetic =~ [aeiou]")
formants <- quantify(vowels, superassp::trk_formant_forest)

# 2. Find outliers
library(dplyr)
outliers <- formants %>%
  group_by(labels) %>%
  mutate(
    F1_z = scale(F1),
    F2_z = scale(F2)
  ) %>%
  filter(abs(F1_z) > 3 | abs(F2_z) > 3) %>%
  ungroup()

cat(sprintf("Found %d outlier measurements\n", nrow(outliers)))

# 3. Review outliers in context
review(corp, outliers)

# Annotator checks:
# - Are boundaries accurate?
# - Is the label correct?
# - Is there background noise?
# - Should this token be excluded?

# 4. Re-run analysis after corrections
vowels_corrected <- query(corp, "Phonetic =~ [aeiou]")
formants_corrected <- quantify(vowels_corrected, superassp::trk_formant_forest)

# 5. Compare results
cat("Before correction:\n")
summary(formants$F1)

cat("\nAfter correction:\n")
summary(formants_corrected$F1)
```

## Collaborative annotation

For team annotation projects:

```{r collaborative}
# 1. Set up bundle lists for each annotator
corp <- demo_corpus()   # stands in for corpus("path/to/database_emuDB")

# Create bundle lists (one-time setup, using emuR)
# emuR::write_bundleList(get_emuDBhandle(corp), "Annotator_A", bundles_A)
# emuR::write_bundleList(get_emuDBhandle(corp), "Annotator_B", bundles_B)

# 2. Each annotator serves their assigned list
annotate(corp, bundleListName = "Annotator_A")

# 3. Track progress
bundles_A <- emuR::read_bundleList(get_emuDBhandle(corp), "Annotator_A")

# Check which have been annotated
# (Look for METADATA.json or specific annotation levels)
```

## Troubleshooting

### Common issues

**Problem:** WebApp doesn't open

```{r troubleshoot-open}
# Check if port is already in use
annotate(corp, port = 8080)  # Try different port

# Check browser setting
annotate(corp, browser = getOption("browser"))
```

**Problem:** "Artic (the annotation web app) was not found"

```{r troubleshoot-path}
# Point at a build explicitly
options(reindeer.artic.dir = "/path/to/artic/dist")

# Inspect what was resolved, and install a build if needed
artic_info()
install_artic(local = "../artic/dist")
```

**Problem:** Changes not saving

```{r troubleshoot-save}
# Ensure database isn't read-only
# Check file permissions on the database directory

# Verify changes are reflected
load_metadata(corp, verbose = TRUE)
```

**Problem:** Server won't stop

```{r troubleshoot-stop}
# Force stop all servers
httpuv::stopAllServers()

# If that doesn't work, restart R session
.rs.restartR()  # RStudio only
```

### Getting help

```{r help}
# Function documentation
?annotate
?review

# Check server status
# (No built-in status check, but server will error if already running)

# View Artic console for errors
# Open browser developer tools (F12)
```

## Tips and best practices

### 1. Start small

```{r tip-small}
# Don't serve the entire corpus at once
# Start with a small session
annotate(corp, sessionPattern = "pilot_session")

# Then scale up as needed
```

### 2. Use query results for targeted annotation

```{r tip-query}
# Much more efficient than browsing all bundles
vowels <- query(corp, "Phonetic =~ [aeiou]")
review(corp, vowels)
```

### 3. Save regularly

```{r tip-save}
# Artic saves when you save the bundle:
# - The save button is enabled by annotate()/review()
# - Switching bundles prompts if there are unsaved edits
# - Check the database after major annotation sessions
# - Consider git commits for version control
```

### 4. Use the RStudio viewer

```{r tip-viewer}
# Default behavior in RStudio
annotate(corp, useViewer = TRUE)

# Benefits:
# - Side-by-side with R console
# - Easy to copy segment IDs or labels
# - Switch between code and annotation
```

### 5. Keyboard shortcuts

Learn the keyboard shortcuts - they dramatically speed up annotation:
- Space: Play/pause
- Arrow keys: Navigate
- Enter: Confirm edit
- Esc: Cancel edit

### 6. Check configuration

```{r tip-config}
# Before a long annotation session:
# 1. Verify correct levels are shown
corp@config$levelDefinitions

# 2. Check label definitions
corp@config$labelDefinitions

# 3. Ensure hierarchy is correct
corp@config$linkDefinitions
```

## Example: complete annotation session

```{r complete-example}
library(reindeer)
library(dplyr)

# === SETUP ===
corp <- corpus("path/to/corpus_emuDB")

# === FIND WORK TO DO ===
# Find all words, check for missing phonetic annotations
words <- query(corp, "Word =~ .*")

# Check which bundles have words but no phonetic tier
bundles_with_words <- words %>%
  select(session, bundle) %>%
  distinct()

phonemes <- query(corp, "Phonetic =~ .*")
bundles_with_phonemes <- phonemes %>%
  select(session, bundle) %>%
  distinct()

bundles_needing_annotation <- anti_join(
  bundles_with_words,
  bundles_with_phonemes,
  by = c("session", "bundle")
)

cat(sprintf("Found %d bundles needing phonetic annotation\n",
            nrow(bundles_needing_annotation)))

# === ANNOTATE ===
# Convert to segment_list format for annotate()
work_segments <- words %>%
  semi_join(bundles_needing_annotation, by = c("session", "bundle"))

# Open annotation interface
review(corp, work_segments)

# [ Annotator works in Artic ]
# - Add phonetic segments within each word
# - Adjust boundaries as needed
# - Save changes (automatic)

# === VERIFY ===
# Check progress after annotation session
phonemes_updated <- query(corp, "Phonetic =~ .*")
bundles_with_phonemes_updated <- phonemes_updated %>%
  select(session, bundle) %>%
  distinct()

completed <- semi_join(
  bundles_needing_annotation,
  bundles_with_phonemes_updated,
  by = c("session", "bundle")
)

cat(sprintf("Completed %d/%d bundles\n",
            nrow(completed),
            nrow(bundles_needing_annotation)))

# === ANALYZE ===
# Now ready for acoustic analysis
phonemes_final <- query(corp, "Phonetic =~ [aeiou]")
formants <- quantify(phonemes_final, superassp::trk_formant_forest)

# Proceed with analysis...
```

## Summary

`annotate()` opens the corpus for editing with one call; `review()` turns a
segment list into a playlist of time anchors. Both filter to exactly the
sessions, bundles, or segments you pass, work directly with query and
quantify results, and save changes back to the database when you save a
bundle.

## See also

- `?annotate`, `?review` - function documentation
- `?install_artic`, `?artic_info` - provision and inspect the web app
- `vignette("getting_started")` - Reindeer basics
- Artic source and releases: https://github.com/humlab-speech/artic
