---
title: "Lazy Pipelines and Pipe-Loss Debugging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lazy Pipelines and Pipe-Loss Debugging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")

# Saving a pipeline with qs2 is shown when the package is available.
eval_qs2 <- requireNamespace("qs2", quietly = TRUE)
```

## Why this vignette exists

Reindeer pipelines are lazy by default: `query()` builds a plan, you
attach `filter()`, `scout()`, `ascend_to()` and so on, and nothing runs
until you ask for data. That is fast — but it also means a filter that
silently drops 80% of your rows can sit unnoticed in the middle of a
long pipe. Reindeer records per-step provenance and warns when a step
drops too many rows.

This vignette shows three things: building and collecting a lazy
pipeline, reading the provenance trail to find where rows disappeared,
and tuning the loss warning to your tolerance.

## A lazy pipeline

```{r plan}
library(reindeer)
library(dplyr, warn.conflicts = FALSE)

corp <- demo_corpus()

plan <- query(corp, "Phoneme =~ .+") |>
  filter(labels != "n") |>
  scout(steps_forward = 1) |>
  ascend_to("Syllable")

class(plan)
```

Nothing has run yet. Printing the plan shows the queued transforms:

```{r print-plan}
plan
```

To materialise, ask for data:

```{r collect}
result <- reindeer::collect(plan)
nrow(result)
head(result) |> select(session, bundle, labels, start, end)
```

dplyr verbs auto-collect, so this works too:

```{r dplyr}
plan |> filter(end - start > 100) |> summarise(n = n())
```

## Reading the provenance trail

Every navigation verb and every dplyr verb appends a row to a provenance
log on the result. Open it with `provenance()`:

```{r provenance}
provenance(result)
```

The same information in row-loss form, friendlier for picking out the
culprit step:

```{r dropped-rows}
dropped_rows(result)
```

`dropped(result)` returns the cumulative count and
`dropped(result, 3)` the loss at a specific step:

```{r dropped}
dropped(result)
```

## The 25 % loss warning

If a navigation verb (`scout()`, `ascend_to()`, `descend_to()`) drops more
than 25 % of its input, reindeer warns by default. Stepping far past the
end of the annotation is the usual cause:

```{r loss-warning}
far <- query(corp, "Phoneme =~ .+") |> scout(steps_forward = 3)
nrow(reindeer::collect(far))

# The warning names the step and the share of rows it lost.
suppressWarnings(provenance(reindeer::collect(far)))
```

dplyr verbs are silent — losing rows there is usually intentional. To
change the threshold:

```{r threshold}
options(reindeer.loss_warn = 0.10)   # warn over 10%
options(reindeer.loss_warn = 1.0)    # never warn
options(reindeer.loss_warn = NULL)   # back to the 25% default
```

The provenance log is capped by `options(reindeer.provenance_max = 1000)`
rows; very long pipelines truncate from the front.

Stepping so far past the end that *nothing* matches is the one case that
still misbehaves: `scout(steps_forward = 99)` aborts inside the
`segment_list` validator instead of returning an empty result. It is
recorded in `dev/REINDEER_REMEDIATION_PLAN.md`.

## Provenance survives serialisation

Save a segment list and its provenance travels with it:

```{r serialise, eval = eval_qs2}
f <- tempfile(fileext = ".qs2")
qs2::qs_save(result, f)
back <- qs2::qs_read(f)
provenance(back) |> nrow()
```

So you can hand a colleague the result of a pipeline and they can still
see how it was filtered.

## Forcing eager mode

For immediate execution, pass `lazy = FALSE` to `query()` or collect
early:

```{r eager}
segs <- reindeer::collect(query(corp, "Phoneme =~ .+"))
nrow(segs)

eager <- query(corp, "Phoneme =~ .+", lazy = FALSE)
class(eager)
```

Eager results carry provenance too, so the inspection workflow is the
same.

## Where next

- `vignette("getting_started")` — the five-minute pipeline.
- `?query`, `?scout`, `?ascend_to`, `?provenance`, `?dropped_rows`.
