---
title: "Lazy Pipelines and Pipe-Loss Debugging"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lazy Pipelines and Pipe-Loss Debugging}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = identical(Sys.getenv("REINDEER_EVAL_VIGNETTES"), "true"))
```

## Why this vignette exists

Reindeer pipelines are lazy by default: `query()` builds a plan, you
attach `filter()`, `scout()`, `ascend_to()`, and so on, and nothing
actually runs until you ask for data. That's fast — but it means a
filter that silently drops 80% of your rows can sit unnoticed in the
middle of a long pipe. Reindeer addresses this by recording per-step
provenance and warning when a step drops too many rows.

This vignette shows three things:

1. Building and collecting a lazy pipeline.
2. Reading the provenance trail to find where rows disappeared.
3. Tuning the loss warning to your tolerance.

## A lazy pipeline

```{r}
library(reindeer)
library(dplyr)

corp <- corpus("path/to/ae_emuDB")

plan <- query(corp, "Phonetic =~ [aeiou]") |>
  filter(label != "@") |>
  scout(steps_forward = 1) |>
  ascend_to("Word")

class(plan)
#> "reindeer::lazy_segment_list"
```

Nothing has run yet. Print the plan and you'll see the queued
transforms with a rough row estimate.

To materialise, ask for data:

```{r}
collect(plan)        # explicit
nrow(plan)           # also triggers
head(plan)           # ditto
plan$labels          # ditto
```

dplyr verbs auto-collect, so this works too:

```{r}
plan |> filter(end - start > 100) |> summarise(n = n())
```

## Reading the provenance trail

Every navigation verb and every dplyr verb appends a row to a
provenance log on the result. Open it with `provenance()`:

```{r}
plan_done <- collect(plan)
provenance(plan_done)
# # A tibble: 4 x 7
#    step verb       call                rows_in rows_out rows_lost ts
#       1 query      query(corp, "...")        NA      812        NA …
#       2 filter     filter(label != "@")     812      798        14 …
#       3 scout      scout(steps_forward=1)   798      798         0 …
#       4 ascend_to  ascend_to("Word")        798      624       174 …
```

The same information in row-loss form, friendlier for picking out the
culprit step:

```{r}
dropped_rows(plan_done)
# step verb       rows_lost pct_lost
#    2 filter            14      1.7
#    4 ascend_to        174     21.8
```

`dropped(plan_done)` returns the cumulative count
(`14 + 174 = 188`). `dropped(plan_done, 4)` returns the loss at a
specific step.

## The 25 % loss warning

If any navigation verb (`scout()`, `ascend_to()`, `descend_to()`)
drops more than 25 % of its input, reindeer warns by default:

```{r}
plan |> scout(steps_forward = 99)
#> Warning: scout dropped 612 rows (76.7%) — most segments hit a
#> bundle boundary at this offset.
```

dplyr verbs are silent — losing rows there is usually intentional. To
change the threshold:

```{r}
options(reindeer.loss_warn = 0.10)   # warn over 10%
options(reindeer.loss_warn = 1.0)    # never warn
```

The cap on log size is governed by `options(reindeer.provenance_max =
1000)` (default 1000 rows). Very long pipelines truncate from the
front.

## Provenance survives serialisation

If you save a segment_list with `saveRDS()` or `qs2::qs_save()`, its
provenance survives the round trip:

```{r}
qs2::qs_save(plan_done, "plan.qs2")
plan_back <- qs2::qs_read("plan.qs2")
provenance(plan_back)        # same tibble as before
```

So you can hand a colleague the result of your pipeline and they can
still see exactly how it was filtered.

## Forcing eager mode

If you want immediate execution — usually for debugging — pass
`lazy = FALSE` to `query()`:

```{r}
query(corp, "Phonetic == t", lazy = FALSE)
```

Or call `collect()` early:

```{r}
segs <- collect(query(corp, "Phonetic == t"))
```

Eager pipelines also record provenance, so the inspection workflow is
identical.

## Where next

- `vignette("getting_started")` — five-minute introduction.
- `?query`, `?scout`, `?ascend_to`, `?provenance`, `?dropped_rows`.
