---
title: "Metadata Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## Why metadata matters in reindeer

Speaker metadata is what lets `quantify()` and `enrich()` pick the right
DSP parameters for each recording — child versus adult, female versus
male, fast versus slow speech. Set the metadata once, then every
downstream analysis uses age- and gender-aware windows, formant ranges
and pitch limits. This vignette covers how to write, read and round-trip
that metadata.

Everything below runs against the demo corpus that ships with the
package.

```{r open}
library(reindeer)
library(dplyr, warn.conflicts = FALSE)

corp <- demo_corpus()
get_metadata(corp) |> head(3)
```

## Three levels, one rule

Metadata can live at three levels:

1. **Database** — defaults that apply to every bundle.
2. **Session** — overrides for one session.
3. **Bundle** — overrides for one recording.

When you ask for a value, **bundle wins over session, session wins over
database**. This lets you set `Project = "MyStudy"` once at the database
level and override only what differs.

## Writing metadata

`set_metadata()` is the canonical writer. It writes to `METADATA.json` on
disk (the ground truth) and refreshes the cache.

```{r write}
# Database-wide defaults
set_metadata(corp, list(
  Project = "Vowel Perception Study",
  Year = 2026,
  Institution = "Humlab"
))

# Session-level
set_metadata(corp, list(Speaker = "P001", Age = 25, Gender = "Female"),
             session = "0000")

# Bundle-level: details for one recording
set_metadata(corp, list(Quality = "Excellent", Microphone = "SM58"),
             session = "0000", bundle = "msajc003")
```

Inheritance in action — the database default shows up everywhere, the
session and bundle values only where they were set:

```{r inherit}
get_metadata(corp) |>
  dplyr::select(session, bundle, Project, Speaker, Age, Gender,
                Quality, Microphone) |>
  utils::head(4)
```

A bracket shortcut writes at the bundle level:

```{r bracket}
md <- get_metadata(corp)
corp[md$session[1], md$bundle[1]] <- list(SamplingRate = 44100)
```

The same bracket reads. `corp[session, bundle]` returns the bundle's
resolved metadata as a list, and you can pass regex on either side to
scan a slice of the corpus:

```{r bracket-read, eval = FALSE}
corp[md$session[1], md$bundle[1]]    # all metadata for one bundle
```

Reading one bundle back through the bracket form currently fails when a field
resolves through more than one level, so the supported read path is used here;
the defect is recorded in `dev/REINDEER_REMEDIATION_PLAN.md`.

```{r bracket-read-alt}
get_metadata(corp) |> utils::head(3)
get_metadata(corp, session_pattern = md$session[1]) |> nrow()
```

Assigning a character path imports a media file at the bundle scope:

```{r bracket-path, eval = FALSE}
corp["0000", "msajc003"] <- "audio/take1.wav"
```

`add_metadata()` is an older spelling of `set_metadata()`; both write the
same way. New code should use `set_metadata()`.

## Reading metadata

`get_metadata()` returns one row per bundle with inheritance already
resolved.

```{r read}
get_metadata(corp) |> nrow()
get_metadata(corp, session_pattern = "0000") |> nrow()
```

For most pipelines you do not need to call this directly — use
`enrich(segs, corp)` and reindeer joins the metadata for you.

## Round-tripping through a spreadsheet

For batch editing, export everything to Excel, edit there, and load it
back:

```{r excel, eval = requireNamespace("openxlsx", quietly = TRUE)}
xlsx <- tempfile(fileext = ".xlsx")
export_metadata(corp, xlsx)

# ... edit in Excel: rows are bundles, columns are fields ...

load_metadata(corp, source = "excel", path = xlsx)
```

The workbook has three sheets — `bundles`, `sessions`, `database` —
matching the inheritance levels. Add columns to introduce new fields;
clear cells to drop them.

## Refreshing after manual edits

If you (or a collaborator) edited the `METADATA.json` files outside R,
tell reindeer to rescan:

```{r refresh}
load_metadata(corp)            # source = "files" is the default
```

`gather_metadata(corp)` does the same thing and is kept for older code.

## Sanity-checking metadata

After a big edit, look at what is there and what `quantify()` will pick:

```{r sanity}
get_metadata(corp) |> glimpse()

dsp_parameters(age = 8,  gender = "Male")     # what params for a kid?
dsp_parameters(age = 45, gender = "Female")   # …for an adult?
```

If the corpus has `Age`/`Gender` on every bundle, `dsp_parameters(corp)`
gives you a per-bundle parameter table — a quick check that no recording
will be analysed with the wrong settings.

## FAIR / CMDI export

`describe_corpus()` writes the archival artefacts next to the corpus:
`README.md`, a CMDI XML file, DataCite JSON, `CITATION.cff` and a
schema.org JSON-LD descriptor. With `options(reindeer.auto_cmdi = TRUE)`
it runs automatically the next time you open the corpus after a metadata
change.

```{r fair}
describe_corpus(corp)
list.files(corp@basePath, pattern = "README|cmdi|datacite|CITATION|jsonld")
```

## Where next

- `vignette("getting_started")` — the five-minute pipeline.
- `vignette("end_to_end_pipeline")` — metadata through to archival output.
- `?set_metadata`, `?get_metadata`, `?load_metadata`, `?dsp_parameters` —
  function reference.
