---
title: "Metadata Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Metadata Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = identical(Sys.getenv("REINDEER_EVAL_VIGNETTES"), "true"))
```

## Why metadata matters in reindeer

Speaker metadata is what lets `quantify()` and `enrich()` pick the
right DSP parameters for each recording — child versus adult, female
versus male, fast versus slow speech. Set the metadata once, then
every downstream analysis automatically uses age- and gender-aware
windows, formant ranges, and pitch limits. This vignette covers how
to write, read, and round-trip that metadata.

## Three levels, one rule

Metadata can live at three levels:

1. **Database** — defaults that apply to every bundle.
2. **Session** — overrides for one session.
3. **Bundle** — overrides for one recording.

When you ask for a value, **bundle wins over session, session wins
over database**. This lets you set `Project = "MyStudy"` once at the
database level and only override what differs.

## Writing metadata

`set_metadata()` is the canonical writer. It writes to
`METADATA.json` on disk (the ground truth) and refreshes the cache.

```{r}
library(reindeer)

corp <- corpus("path/to/ae_emuDB")

# Database-wide defaults
set_metadata(corp, list(
  Project = "Vowel Perception Study",
  Year = 2026,
  Institution = "Humlab"
))

# Session-level: one speaker for all that session's bundles
set_metadata(corp,
             list(Speaker = "P001", Age = 25, Gender = "Female"),
             session = "Session1")

# Bundle-level: details for one recording
set_metadata(corp,
             list(Quality = "Excellent", Microphone = "SM58"),
             session = "Session1", bundle = "Recording001")
```

A bracket shortcut writes at the bundle level:

```{r}
corp["Session1", "Recording001"] <- list(SamplingRate = 44100)
```

The same bracket also reads — `corp[session, bundle]` returns the
bundle's resolved metadata as a list, and you can pass regex on either
side to scan a slice of the corpus:

```{r}
corp["Session1", "Recording001"]    # all metadata for one bundle
corp["Session1", ]                  # every bundle in Session1
corp["Sess.*", "Recording0[1-3].*"] # regex on both axes
```

Assigning a character path imports a media file at the bundle scope:

```{r}
corp["Session1", "Recording001"] <- "audio/take1.wav"
```

`add_metadata()` is a deprecated alias of `set_metadata()`; existing
scripts keep working, but new code should use `set_metadata()`.

## Reading metadata

`get_metadata()` returns one row per bundle with inheritance already
resolved.

```{r}
get_metadata(corp)                              # everything
get_metadata(corp, session_pattern = "Session1")
get_metadata(corp, bundle_pattern = "Recording0[1-3].*")
```

For most pipelines you don't need to call this directly — use
`enrich(segs, corp)` and reindeer joins the metadata for you.

## Round-tripping through a spreadsheet

For batch editing, export everything to Excel, edit there, and load it
back:

```{r}
export_metadata(corp, "ae_metadata.xlsx")

# ... edit in Excel: rows are bundles, columns are fields ...

load_metadata(corp, source = "excel", path = "ae_metadata.xlsx")
```

The workbook has three sheets — `bundles`, `sessions`, `database` —
matching the inheritance levels. Add columns to introduce new fields;
clear cells to drop them.

## Refreshing after manual edits

If you (or a collaborator) edited the `METADATA.json` files outside R,
tell reindeer to rescan:

```{r}
load_metadata(corp)            # source = "files" is the default
```

`gather_metadata(corp)` is a deprecated alias and does the same thing.

## Sanity-checking metadata

After a big edit, look at what's there and what `quantify()` will pick:

```{r}
get_metadata(corp) |> dplyr::glimpse()        # column overview

dsp_parameters(age = 8,  gender = "Male")     # what params for a kid?
dsp_parameters(age = 45, gender = "Female")   # …for an adult?
dsp_parameters(corpus_obj = corp)             # one row per bundle
```

If the corpus has Age/Gender on every bundle, `dsp_parameters(corp)`
gives you a per-bundle parameter table — a quick check that no
recording will be analysed with the wrong settings.

## FAIR / CMDI export

`describe_corpus()` writes README.md, a CMDI XML file, and DataCite
JSON next to the corpus. With `options(reindeer.auto_cmdi = TRUE)` it
runs automatically the next time you open the corpus after any
metadata change.

```{r}
describe_corpus(corp)
options(reindeer.auto_cmdi = TRUE)
```

## Where next

- `vignette("getting_started")` — the five-minute pipeline.
- `?set_metadata`, `?get_metadata`, `?load_metadata`,
  `?dsp_parameters` — function reference.
