---
title: "Automatic Transcription with reindeeR"
author: "reindeeR Team"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic Transcription with reindeeR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)

library(reindeeR)
```

## Introduction

The reindeeR transcription system provides a comprehensive workflow for integrating automatic annotation (e.g., from ASR systems) into Emu-SDMS databases. The workflow consists of five key stages:

1. **Draft**: Create transcription suggestions from automatic annotation
2. **Assess**: Validate suggestions against database constraints
3. **Correct**: Manually adjust specific suggestions
4. **Prepare**: Create/update levels and label types as needed
5. **Transcribe**: Apply suggestions to database with logging

This approach ensures data integrity while providing flexibility for manual correction and quality control.

## Basic Workflow

### Step 1: Load Database as Corpus

```{r load-corpus, eval=FALSE}
library(reindeeR)

# Load an Emu-SDMS database as a corpus object
my_corpus <- corpus("path/to/database_emuDB")
```

### Step 2: Draft Transcription Suggestions

The `draft()` function takes a corpus and an annotation function that generates automatic transcriptions:

```{r draft-example, eval=FALSE}
# Define an annotation function
# This could call an ASR API, run a local model, etc.
my_annotator <- function(corpus, session, bundle, ...) {
  # Your automatic annotation logic here
  # Must return a data.frame with columns: start_time, end_time, label
  
  # Example: Mock annotation output
  data.frame(
    start_time = c(0.0, 1.2, 2.5),
    end_time = c(1.0, 2.3, 3.8),
    label = c("hello", "world", "speech")
  )
}

# Create suggestions
suggestions <- draft(
  corpus = my_corpus,
  annotation_func = my_annotator,
  session = "session001",
  bundle = "bundle001",
  level_name = "Word",
  level_type = "ITEM"
)

# View suggestions
print(suggestions)
```

### Step 3: Assess Suggestions

Before applying suggestions, validate them against database constraints:

```{r assess-example, eval=FALSE}
# Run comprehensive assessment
suggestions <- assess(suggestions, verbose = TRUE)

# The assessment checks:
# - Level and attribute existence
# - Temporal overlap detection
# - Duration constraints
# - Bundle timing boundaries
# - Label validity (if defined in database)
```

### Step 4: Correct Suggestions (Optional)

Manually adjust any problematic suggestions:

```{r correct-example, eval=FALSE}
# Correct the first suggestion
suggestions <- correct(
  suggestions,
  index = 1,
  start_time = 0.05,  # Adjust start time
  label = "Hello"     # Capitalize label
)

# Correct multiple suggestions
suggestions <- correct(
  suggestions,
  index = c(2, 3),
  label = c("World", "Speech")
)

# Reassess after corrections
suggestions <- assess(suggestions, verbose = TRUE)
```

### Step 5: Prepare Database

If the suggestions require new levels or attributes, create them:

```{r prepare-example, eval=FALSE}
# Create required levels/attributes
# This updates the database configuration and rewrites annotation files
updated_config <- prepare(suggestions, force = FALSE, verbose = TRUE)

# force=FALSE will prompt for confirmation
# force=TRUE will proceed without prompting
```

### Step 6: Transcribe

Apply the suggestions to the database and create a log:

```{r transcribe-example, eval=FALSE}
# Apply suggestions
log <- transcribe(suggestions, force = FALSE, verbose = TRUE)

# View log summary
summary(log)

# The log tracks:
# - Items added/removed
# - Levels/attributes created
# - Success/failure status
# - Error messages (if any)
```

## Advanced Features

### Working with Different Level Types

#### Segments

Segments have explicit start and end times and durations:

```{r segments-example, eval=FALSE}
phoneme_suggestions <- SuggestedSegments(
  corpus = my_corpus,
  session = "session001",
  bundle = "bundle001",
  level_name = "Phoneme",
  suggestions = data.frame(
    start_time = c(0.0, 0.1, 0.2),
    end_time = c(0.1, 0.2, 0.3),
    label = c("p", "ɑ", "t")
  ),
  phonetic_alphabet = "IPA",
  min_duration = 0.01  # Minimum 10ms
)
```

#### Events

Events are time points without duration:

```{r events-example, eval=FALSE}
tone_suggestions <- SuggestedEvents(
  corpus = my_corpus,
  session = "session001",
  bundle = "bundle001",
  level_name = "Tone",
  suggestions = data.frame(
    start_time = c(0.5, 1.2, 2.0),
    end_time = c(0.5, 1.2, 2.0),  # Same as start for events
    label = c("H*", "L-", "H%")
  ),
  event_categories = c("pitch_accent", "phrase_accent", "boundary_tone")
)
```

#### Items

Items can have associated confidence scores:

```{r items-example, eval=FALSE}
word_suggestions <- SuggestedItems(
  corpus = my_corpus,
  session = "session001",
  bundle = "bundle001",
  level_name = "Word",
  suggestions = data.frame(
    start_time = c(0.0, 1.0, 2.0),
    end_time = c(0.9, 1.9, 2.9),
    label = c("first", "second", "third")
  ),
  confidence_scores = c(0.95, 0.88, 0.92)
)
```

### Reversing Transcriptions

If you need to undo a transcription:

```{r reverse-example, eval=FALSE}
# Reverse the changes
reverse(log, verbose = TRUE)

# This removes all items added by the transcription
# and restores the annotation file to its previous state
```

### Batch Processing

Process multiple bundles:

```{r batch-example, eval=FALSE}
# Get all bundles in a session
bundles <- my_corpus["session.*", ".*"]

# Process each bundle
logs <- list()
for (i in 1:nrow(bundles)) {
  session <- bundles$session_name[i]
  bundle <- bundles$bundle_name[i]
  
  # Draft
  sugg <- draft(
    corpus = my_corpus,
    annotation_func = my_annotator,
    session = session,
    bundle = bundle,
    level_name = "Word",
    level_type = "ITEM"
  )
  
  # Assess
  sugg <- assess(sugg, verbose = FALSE)
  
  # Only transcribe if no errors
  if (length(sugg@assessment_results$errors) == 0) {
    logs[[i]] <- transcribe(sugg, verbose = FALSE)
  }
}

# Check results
success_count <- sum(sapply(logs, function(x) x@success))
cat(sprintf("Successfully transcribed %d/%d bundles\n", success_count, length(logs)))
```

## Quality Control

### Inspecting Suggestions

```{r inspect-example, eval=FALSE}
# View suggestion data
head(suggestions@suggestions)

# Check specific fields
suggestions@session
suggestions@bundle
suggestions@level_name
suggestions@level_type

# View assessment results
suggestions@assessment_results$errors
suggestions@assessment_results$warnings
suggestions@assessment_results$info
```

### Filtering by Confidence

```{r filter-confidence, eval=FALSE}
# For items with confidence scores
high_confidence <- word_suggestions@suggestions[
  word_suggestions@suggestions$confidence > 0.9,
]

# Create new suggestion object with filtered data
filtered_suggestions <- SuggestedItems(
  corpus = my_corpus,
  session = "session001",
  bundle = "bundle001",
  level_name = "Word",
  suggestions = high_confidence,
  confidence_scores = high_confidence$confidence
)
```

### Duration Filtering

```{r filter-duration, eval=FALSE}
# Set a higher minimum duration
phoneme_suggestions <- SuggestedSegments(
  corpus = my_corpus,
  session = "session001",
  bundle = "bundle001",
  level_name = "Phoneme",
  suggestions = phoneme_data,
  min_duration = 0.02  # 20ms minimum
)

# Suggestions shorter than this will be filtered during transcription
```

## Error Handling

The transcription system provides detailed error messages:

```{r error-handling, eval=FALSE}
# Assessment errors prevent transcription
suggestions <- assess(suggestions)

if (length(suggestions@assessment_results$errors) > 0) {
  cat("Errors found:\n")
  for (err in suggestions@assessment_results$errors) {
    cat("  -", err, "\n")
  }
  
  # Fix errors before proceeding
  # e.g., create missing levels with prepare()
}

# Transcription errors are logged
log <- transcribe(suggestions)

if (!log@success) {
  cat("Transcription failed:", log@error_message, "\n")
}
```

## Integration with ASR Systems

### Example: Whisper Integration

```{r whisper-example, eval=FALSE}
library(reticulate)

# Initialize Whisper (requires Python setup)
whisper <- import("whisper")
model <- whisper$load_model("base")

# Create annotation function
whisper_annotator <- function(corpus, session, bundle, language = "en") {
  # Get audio file path
  audio_file <- signal_files(corpus) %>%
    filter(session == !!session, bundle == !!bundle) %>%
    pull(full_path) %>%
    first()
  
  # Run Whisper
  result <- model$transcribe(
    audio_file,
    language = language,
    word_timestamps = TRUE
  )
  
  # Convert to reindeeR format
  segments <- result$segments
  
  data.frame(
    start_time = sapply(segments, function(s) s$start),
    end_time = sapply(segments, function(s) s$end),
    label = sapply(segments, function(s) s$text)
  )
}

# Use with draft()
suggestions <- draft(
  corpus = my_corpus,
  annotation_func = whisper_annotator,
  session = "session001",
  bundle = "bundle001",
  level_name = "Word",
  level_type = "ITEM",
  language = "en"
)
```

### Example: Montreal Forced Aligner

```{r mfa-example, eval=FALSE}
# Assume MFA outputs TextGrid files
library(rPraat)

mfa_annotator <- function(corpus, session, bundle, textgrid_path) {
  # Read TextGrid
  tg <- tg.read(textgrid_path)
  
  # Extract tier
  tier <- tg.getTier(tg, "phones")
  
  # Convert to data.frame
  data.frame(
    start_time = tier$t1,
    end_time = tier$t2,
    label = tier$label
  )
}

# Use with draft()
suggestions <- draft(
  corpus = my_corpus,
  annotation_func = mfa_annotator,
  session = "session001",
  bundle = "bundle001",
  level_name = "Phoneme",
  level_type = "SEGMENT",
  textgrid_path = "path/to/alignment.TextGrid"
)
```

## Summary

The reindeeR transcription system provides:

- **Type safety**: S7 classes ensure consistent data structures
- **Validation**: Comprehensive checks before applying changes
- **Flexibility**: Manual correction of automatic annotations
- **Logging**: Track all changes for reproducibility and rollback
- **Integration**: Easy connection to any annotation tool or API

This workflow bridges the gap between automatic annotation and curated linguistic databases, making it practical to leverage modern speech technologies while maintaining data quality.

## See Also

- `?corpus` - Corpus object documentation
- `?draft` - Create transcription suggestions
- `?assess` - Validate suggestions
- `?transcribe` - Apply suggestions to database
- `vignette("query_benchmarks")` - Query optimization benchmarks
