---
title: "Automatic Transcription (moved to protoscribe)"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Automatic Transcription (moved to protoscribe)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Transcription Workflow Has Moved

As of reindeer v0.3.0, all automatic transcription and draft annotation
functionality has been moved to the companion package
**[protoscribe](https://github.com/humlab-speech/protoscribe)**.

This includes the complete draft-assess-prepare-transcribe-reverse workflow,
all suggestion classes (`SuggestedSegments`, `SuggestedEvents`, `SuggestedItems`),
and the draft cache system.

## Quick Start with protoscribe

```{r protoscribe-example, eval=FALSE}
library(reindeer)
library(protoscribe)

# Load corpus with reindeer
corp <- corpus("path/to/database_emuDB")

# Use protoscribe for annotation drafting
# See protoscribe documentation for available draft_*() functions
suggestions <- protoscribe::draft_vad(corpus = corp)

# Validate suggestions
suggestions <- protoscribe::assess(suggestions)

# Apply to database
protoscribe::transcribe(suggestions)
```

## See Also

- [protoscribe on GitHub](https://github.com/humlab-speech/protoscribe)
- `vignette("getting_started")` -- reindeer basics
- `?corpus` -- Corpus object documentation
- `?query` -- Query segments from a corpus
